---
title: "Notes on Chapter 26 - Iteration"
author: "The Caveman Coder"
date: "2025-08-16"
output:
  pdf_document: default
  html_document: default
---

This chapter introduces the concept of **functional programming**. 


### Prerequisites
```{r}
library(tidyverse)
```

Situation: We are tasked to count the number of observations and compute the median in the tibble below:
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

This can be solved using copy-paste:
```{r}
df |> summarize(
  n = n(),
  a = median(a),
  b = median(b),
  c = median(c),
  d = median(d)
)
```

This breaks the rule of never copy-pasting more than twice, and is prone to mistakes.  

Using `across()` to solve the problem above:
```{r}
df |> summarize(
  n = n(),
  across(a:d, median)
)
```

Voila!


### Selecting columnms with `.cols`

- As the first argument to `across()`, `.cols` specifies which columns you want to iterate over.  
- The second argument `.fns`specifies what to do with each column. 

Example using `across()` with `everything()`:
```{r}
df <- tibble(
  grp = sample(2, 10, replace = TRUE),
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df |> 
  group_by(grp) |> 
  summarize(across(everything(), median))
```


`where()` allows the user to select columns based on type:  

- `where(is.numeric)` selects numeric columns.  
- `where(is.character)` selects all string columns.  
- `where(is.Date)` selects all date columns.  
- `where(is.POSIXct)` selects all date-time columns.  
- `where(is.logical)` selects all logical columns.  


### Calling a single function

Remember, do not put `()` after the function that you are passing to `across()`. The code below will throw and error:
```r
df |> 
  group_by(grp) |> 
  summarize(across(everything(), median()))
```

What if we want to pass an argument to the function we are passing to `across()`? For example, the argument `na.rm = TRUE` on `median()`.

Solution: instead of calling the target function directly, we create
a new function that calls our target function.

Example:
```{r}
rnorm_na <- function(n, n_na, mean = 0, sd = 1) {
  sample(c(rnorm(n - n_na, mean = mean, sd = sd), rep(NA, n_na)))
}

df_miss <- tibble(
  a = rnorm_na(5, 1),
  b = rnorm_na(5, 1),
  c = rnorm_na(5, 2),
  d = rnorm(5)
)

# This is the solution
df_miss |> 
  summarize(
    across(a:d, function(x) median(x, na.rm = TRUE)),
    n = n()
  )
```


There is shorcut for this "throwaway" function: the `\`. Just replace the word `function` with `\`.

Example of handling the anoymous or throwaway function using `\`:
```{r}
df_miss |> 
  summarize(
    across(a:d, \(x) median(x, na.rm = TRUE)),
    n = n()
  )
```


How to know how many values were removed when we specified `na.rm = TRUE`? By supplying another function to `across()`:
```{r}
df_miss |> 
  summarize(
    across(a:d, list(
      median = \(x) median(x, na.rm = TRUE),
      n_miss = \(x) sum(is.na(x))
    )),
    n = n()
  )
```


### Column names

Specifying what to name the results using the `.names` argument:
```{r}
df_miss |> 
  summarize(
    across(
      a:d,
      list(
        median = \(x) median(x, na.rm = TRUE),
        n_miss = \(x) sum(is.na(x))
      ),
      .names = "{.fn}_{.col}"
    ),
    n = n()
  )
```


When using `across()` with `mutate()`, the default action is for `across()` to use the same names as inputs. This means that if you don't use the `.names` argument, the original values of the columns will be replaced. 

Example using `coalesce()` to replace `NA`s with `0`:
```{r}
df_miss |> 
  mutate(
    across(a:d, \(x) coalesce(x, 0))
  )
```

If this is not your intended action, and would instead prefer to create new columns, we can specify the naming format for the new names using `.names`:
```{r}
df_miss |> 
  mutate(
    across(a:d, \(x) coalesce(x, 0), .names = "{.col}_na_zero")
  )
```


### Filtering

For filtering, `if_any()` and `if_all()` are the variants of `across()` provided by dplyr.  

Example:
```{r}
df_miss |> filter(if_any(a:d, is.na))
df_miss |> filter(if_all(a:d, is.na))
```


### `across()` in functions

Date expander from Jacob Scott:
```{r}
expand_dates <- function(df) {
  df |> 
    mutate(
      across(where(is.Date), list(
        year = year,
        month = month,
        dat = mday
      ))
    )
}

df_date <- tibble(
  name = c("Amy", "Bob"),
  date = ymd(c("2009-08-03", "2010-01-16"))
)

df_date |> 
  expand_dates()
```


Example computing the means of numeric columns:
```{r}
summarize_means <- function(df, summary_vars = where(is.numeric)) {
  df |> 
    summarize(
      across({{ summary_vars }}, \(x) mean(x, na.rm = TRUE)),
      n = n(),
      .groups = "drop"
    )
}

diamonds |> 
  group_by(cut) |> 
  summarize_means()

diamonds |> 
  group_by(cut) |> 
  summarize_means(c(carat, x:z))
```


### Compare with `pivot_longer()`

Example of code using `across()` and `summarize()` to do "pivot" operations:
```{r}
df |> 
  summarize(across(a:d, list(median = median, mean = mean)))
```


We could compute the same values using `pivot_longer()` and `summarize()`:
```{r}
long <- df |> 
  pivot_longer(a:d) |> 
  group_by(name) |> 
  summarize(
    median = median(value),
    mean = mean(value)
  )

long
```

To make the format similar to the first result, we could pivot again:
```{r}
long |> 
  pivot_wider(
    names_from = name,
    values_from = c(median, mean),
    names_vary = "slowest",
    names_glue = "{name}_{.value}"
  )
```

Using this technique will enable us to do more complicated operations that is not directly possible with `across()`. For example, consider the data below:
```{r}
df_paired <- tibble(
  a_val = rnorm(10),
  a_wts = runif(10),
  b_val = rnorm(10),
  b_wts = runif(10),
  c_val = rnorm(10),
  c_wts = runif(10),
  d_val = rnorm(10),
  d_wts = runif(10)
)

df_paired
```

Suppose we want to compute the weighted means, there is currently no way to do this with `across()`, but this could be solved with `pivot_longer()`:
```{r}
df_long <- df_paired |> 
  pivot_longer(
    everything(),
    names_to = c("group", ".value"),
    names_sep = "_"
  )

df_long

df_long |> 
  group_by(group) |> 
  summarize(mean = weighted.mean(val, wts))
```

Similar to the example before, we could use `pivot_wider()` if needed.


### Exercises

xxxxxxx



## Reading multiple files

### Listing files in a directory

Syntax is `list.files(path, pattern, full.names = TRUE)` where:  
- `path` is the directory to look in.  
- `pattern` is the regex used to filter the names. The most common pattern is `[.]xlsx$` or `[.]csv$` to find the files by specifying the extension name.  
- if `full.names` is set to `TRUE`, the full directory name will be included in the output. 

Example:

Listing all Excel files in the specified directory:
```{r}
paths <- list.files("data/gapminder", pattern = "[.]xlsx$", full.names = TRUE)

paths
```


### Lists

Creating a list of Excel files by reading the files specified in the path names:
```{r}
files <- list(
  readxl::read_excel("data/gapminder/1952.xlsx"),
  readxl::read_excel("data/gapminder/1957.xlsx"),
  readxl::read_excel("data/gapminder/1962.xlsx"),
  readxl::read_excel("data/gapminder/1967.xlsx"),
  readxl::read_excel("data/gapminder/1972.xlsx"),
  readxl::read_excel("data/gapminder/1977.xlsx"),
  readxl::read_excel("data/gapminder/1982.xlsx"),
  readxl::read_excel("data/gapminder/1987.xlsx"),
  readxl::read_excel("data/gapminder/1992.xlsx"),
  readxl::read_excel("data/gapminder/1992.xlsx"),
  readxl::read_excel("data/gapminder/2002.xlsx"),
  readxl::read_excel("data/gapminder/2007.xlsx")
)
```


Accessing the target file from the list of files:
```{r}
files[[3]]
```


### `purrr::map()` and `list_rbind()`

If collecting the files by hand is tedious, this method provides a smarter way of doing things:

Using `map()` to get a list of data frames:
```{r}
files <- map(paths, readxl::read_excel)
length(files)

# test
files[[3]]
```


Combining all the files into a single file using `purrr::list_rbind()`: 
```{r}
list_rbind(files)
```


Doing the steps above in a pipeline:
```{r}
paths |> 
  map(readxl::read_excel) |> 
  list_rbind()
```

In case it is necessary to pass more arguments to `read_excel()`, the technique for handling anonymous functions (aka lambda functions) can also be used here:
```{r}
paths |> 
  map(\(path) readxl::read_excel(path, n_max = 1)) |> 
  list_rbind()
```


### Data in the path 

Extracting the filename from the full path using `set_names(basename)`:
```{r}
paths |> set_names(basename)
```

These set names are automatically carried along by all the map functions, allowing us to do this:
```{r}
files <- paths |> 
  set_names(basename) |> 
  map(readxl::read_excel)

files[["1962.xlsx"]]
```

Writing the filenames into a new column during `rbind`:
```{r}
paths |> 
  set_names(basename) |> 
  map(readxl::read_excel) |> 
  list_rbind(names_to = "source")
```

If the filepath contains important information, we can include them in the data like this:
```{r}
paths |> 
  set_names() |> 
  map(readxl::read_excel) |> 
  list_rbind(names_to = "source") |> 
  separate_wider_delim(source, delim = "/", names = c(NA, "dir", "file")) |> 
  separate_wider_delim(file, delim = ".", names = c("name", "ext"))
```


### Saving

```{r}
gapminder <- paths |> 
  set_names() |> 
  map(readxl::read_excel) |> 
  list_rbind(names_to = "source") |> 
  separate_wider_delim(source, delim = "/", names = c(NA, "dir", "file")) |> 
  separate_wider_delim(file, delim = ".", names = c("name", "ext"))

write_csv(gapminder, "data/gapminder.csv")
```


### Many simple iterations

Example of a function that reads in data from multiple files, filters out missing values, pivot, and the combines all the data:
```r
process_file <- function(path) {
  df <- read_csv(path)
  
  df |> 
    filter(!is.na(id)) |> 
    mutate(id = tolower(id)) |> 
    pivot_longer(jan:dec, names_to = "month")
}

paths |> 
  map(process_file) |> 
  list_rbind()

```

Alternative approach: perform each step of `process_file()` to every file:
```r
paths |> 
  map(read_csv) |> 
  map(\(df) df |> filter(!is.na(id))) |> 
  map(\(df) df |> mutate(id = tolower(id))) |> 
  map(\(df) df |> pivot_longer(jan:dec, names_to = "month)) |> 
  list_rbind()
```
The second approach is better because it stops the user from getting fixated on getting the first file right before moving on to the rest. 

Another approach is to bind all the data frames together before doing the other operations:
```r
paths |> 
  map(read_csv) |> 
  list_rbind() |> 
  filter(!is.na(id)) |> 
  mutate(id = tolower(id)) |> 
  pivot_longer(jan:dec, names_to = "month")
```


### Heterogeneous data

Sometimes, it is not possible to go from `map()` straight to `list_rbind()` because the data frames are so heterogeneous that `list_rbind()` either fails or yields a data frame that is not very useful.  

A solution to problems like this usually involves loading all the files first and then capturing the structure of the data frames when you explore the resulting data:
```{r}
df_types <- function(df) {
  tibble(
    col_name = names(df),
    col_type = map_chr(df, vctrs::vec_ptype_full),
    n_miss = map_int(df, \(x) sum(is.na(x)))
  )
}

df_types(gapminder)
```


Then the function can be applied to all the files:
```{r}
files |> 
  map(df_types) |> 
  list_rbind(names_to = "file_name") |> 
  select(-n_miss) |> 
  pivot_wider(names_from = col_name, values_from = col_type)
```


If the files have heterogeneous formats, you might need to do more processing before you can successfully merge them. 


### Handling failures

One of the downsides of `map()`: it either succeeds of fails as a whole.  Solution: The purr package called `possibly()`.

Using `possibly()` to tackle an erring function:
```{r}
files <- paths |> 
  map(possibly(\(path) readxl::read_excel(path), NULL))

# testing
data <- files |> list_rbind()

data

# trying the function on something that will fail
failed <- map_vec(files, is.null)
paths[failed]
```


The function can be called again for each dailure and figure out wha went wrong.  


## Saving multiple outputs

How to take one or more R objects and save it to one or more files?


### Writing to a database


#### Building the data "by hand"

Creating a template by reading a single file and adding more files to it:
```{r}
template <- readxl::read_excel(paths[1])
template$year <- 1952
template
```

Using `DBI::diCreateTable()` to turn the template into a database table:
```{r}
con <- DBI::dbConnect(duckdb::duckdb())
DBI::dbCreateTable(con, "gapminder", template)
```

Inspecting the newly-created db table:
```{r}
con |> tbl("gapminder")
```

Creating a function that takes a single filepath, reads it into R, and then adds the result to the `gapminder` table. 
```{r}
append_file <- function(path) {
  df <- readxl::read_excel(path)
  df$year <- parse_number(basename(path))
  
  DBI::dbAppendTable(con, "gapminder", df)
}
```


One possible code for appending the files for each filepath in `path`:
```r
paths |> map(append_file)
```

Better way: Use `walk()` instead of `map()`. The latter does exactly the same thing as `map()` but throws the ouput away:
```{r}
paths |> walk(append_file)
```


Checking to see if all the data are appended to the table:
```{r}
con |> 
  tbl("gapminder") |> 
  count(year)
```


### Writing `.csv` files

Imagine that we want to take the `ggplot2::diamonds` data and save one csv file for each clarity. 

There are many ways to skin the cat, but the authors recommend using `group_nest()`:
```{r}
by_clarity <- diamonds |> 
  group_nest(clarity)

by_clarity
```

The `data` column is a list-column that contains one tibble for each unique value of `clarity`:
```{r}
by_clarity$data[[1]]
```

Adding a new column to `by_clarity`, to inform us of which filename to assign to each tibble:
```{r}
by_clarity <- by_clarity |> 
  mutate(path = str_glue("diamonds-{clarity}.csv"))

by_clarity
```

In saving the files "by hand," we need to write something like this:
```r
write_csv(by_clarity$data[[1]], by_clarity$path[[1]])
write_csv(by_clarity$data[[2]], by_clarity$path[[2]])
...
```

Since there are two arguments that are changing here, the function `map()` will not work here. Instead, we can use `map2()`. However, because we don't care to see the output, it is better to use `walk2()`. 
```{r}
walk2(by_clarity$data,by_clarity$path, write_csv)
```


### Saving plots

The same approach can be used to create many plots -- just write the correct function:

```{r}
carat_histogram <- function(df) {
  ggplot(df, aes(x = carat)) +
    geom_histogram(binwidth = 0.1)
}

carat_histogram(by_clarity$data[[1]])
```

Now we can use `map()` to create a list of many plots and their eventual filepaths:
```{r}
by_clarity <- by_clarity |> 
  mutate(
    plot = map(data, carat_histogram),
    path = str_glue("clarity-{clarity}.png")
  )

by_clarity
```

Finally, we can use `walk2()` again, this time, with `ggsave()` to save each plot:
```{r}
walk2(
  by_clarity$path,
  by_clarity$plot,
  \(path, plot) ggsave(path, plot, width = 6, height = 6)
)
```







