---
title: 'Notes on Ch 27: A Field Guide to Base R'
author: "The Caveman Coder"
date: "2025-08-18"
output:
  pdf_document: default
  html_document: default
---

Topics covered:  

- Subsetting with `[]`  
- Subsetting with `[[]]`
- Subsetting with `$`
- `For` loops
- Essential plotting functions


### Prerequisites
```{r}
library(tidyverse)
```


## Selecting multiple elements with `[]`

### Subsetting vectors
```{r}
# Accessing elements by position
x <- c("one", "two", "three", "four", "five")

x[c(3, 2, 5)]
```


Since you can repeatedly access a postion, the term "subsetting" is actually a misnomer:
```{r}
x[c(1, 1, 5, 5, 5, 2)]
```

Dropping the elements at specified positions using negative index positions:
```{r}
x[c(-1, -3, -5)]
```

Subsetting with a logical vector:
```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)


# All non-missing values of x
x[!is.na(x)]

# All even (or missing!) values of x
x[x %% 2 == 0]
```

Unlike `filter()`, `NA` indices will be included in the output as `NA`s.

Subsetting with a character vector:
```{r}
x <- c(abc = 1, def = 2, xyz = 5)
x[c("xyz", "def")]
```


Subsetting with "nothing" returns the entire vector:
```{r}
x[]
```


This is a feature, not a bug. Useful when subsetting 2d structures like tibbles.


### Subsetting data frames

Format is `df[rows, columns]`. 

Examples:
```{r}
df <- tibble(
  x = 1:3,
  y = c("a", "e", "f"),
  z = runif(3)
)

# Selecting the first row and second column
df[1, 2]

# Selecting all rows and columns x and y
df[, c("x", "y")]

# Selecting rows where x is greater than 1 and all columns
df[df$x > 1, ]
```

The dollar sign `$` is used to access the column name.

In many applications, a tibble and data frame can be used interchangeably, but in base R, it is important to make a distinction between them: If the target object is a data frame, subsetting a single column with `[]` returns a vector, and subsetting multiple columns with `[]` returns a data frame. If the target object is a tibble, subsetting with `[]` will always return a tibble.
```{r}
# Dataframe
df1 <- data.frame(x = 1:3)
df1[, "x"]

# Tibble
df2 <- tibble(x = 1:3)
df2[, "x"]
```


To avoid this ambiguity, we can use the argument `drop = FALSE` with dataframes:
```{r}
df1[, "x", drop = FALSE]
```


### `dplyr` equivalents

The dplyr function `filter()` is equivalent to subsetting rows with a logical vector, with the missing values excluded:
```{r}
df <- tibble(
  x = c(2, 3, 1, 1, NA),
  y = letters[1:5],
  z = runif(5)
)

df |> filter(x > 1)

# same as
df[!is.na(df$x) & df$x > 1, ]

# same as
df[which(df$x > 1), ]
```


The dplyr function `arrange()` is equivalent to subsetting the rows with an integer vector created with `order()`:
```{r}
df |> arrange(x, y)

# same as
df[order(df$x, df$y), ]
```


Both dplyr functions `select()` and `relocate()` are similar to subsetting the columns with a character vector:
```{r}
df |> select(x, z)

# same as
df[, c("x", "z")]
```

The base R function `subset()` combines the features of the dplyr functions `filter()` and `select()`:
```{r}
df |> 
  filter(x > 1) |> 
  select(y, z)

# same as
df |> subset(x > 1, c(y, z))
```


### Exercises

1. Create functions that take a vector as input and return:

a. The elements at even-numbered positions. 
b. Every element except the last value.  
c. Only even values (and no missing value).  

2. Why is `x[-which(x > 0)]` not the same as `x[x <= 0]`? Read the documentation for `which()` and do some experiments to figure it out. 


## Selecting a single element with `$` and `[[]]`

These two are used to extract a single element. 

### Data frames

`$` is specialized to access by column name, whil `[[]]` is used to access by position or by name:
```{r}
tb <- tibble(
  x = 1:4,
  y = c(10, 4, 1, 21)
)

# by position
tb[[1]]

# by name
tb[["x"]]
tb$x
```


Both can also be used to create new columnms -- the base R equivalent to `mutate()`:
```{r}
tb$z <- tb$x + tb$y
tb
```


The `$` accessor is very convenient to use when performing quick summaries:
```{r}
max(diamonds$carat)
levels(diamonds$cut)
```


In dplyr, `pull()` takes a variable name or variable position and returns just that column, and can be equivalent to the use of accessors in the code above:
```{r}
diamonds |> pull(carat) |> max()

diamonds |> pull(cut) |> levels()
```

### Tibbles

The dollar `$` accessor performs partial matching when used on dataframes:
```{r}
df <- data.frame(x1 = 1)
df$x
df$z
```

When used in tibbles, the `$` acessor is more strict:
```{r}
tb <- tibble(x1 = 1)

tb$x
tb$z
```


Here is a"joke" to remember this distinction between tibbles and dataframes:

> Tibbles are lazy and surly: they do less and complain more.  


### Lists

The `$` and `[[]]` "accessors extract a single component from a list:
```{r}
l <- list(
  a = 1:3,
  b = "a string",
  c = pi,
  d = list(-1, -5)
)

str(l[1:2])
str(l[[1]])

str(l[1])
str(l[4])
str(l[[4]])

str(l$a)
```

Consider this analogy: A pepper shaker that contains sachets of ground pepper.  

- `pepper[1]` refers to a pepper shaker containing a single sachet. 
- `pepper[2]` refers to the same pepper shaker containing the second sachet. 
- `pepper[1:2]`refers to the same pepper shaker containing the first and second sachet.  
- `pepper[[1]]` extracts the first pepper sachet itself.  


## Exercises

1. What happens when you use `[[]]` with a positive integer that's bigger than the length of the vector? What happens when you subset with a name that doesn't exist?  

2. What would `pepper[[1]][1]` be? What about `pepper[[1]][[1]]?


## Apply family 

- The most important function in this family is the `lapply()` function, which can be used to replace every `map()` function, if desired.  
- There is no base R equivalent to `across()`, but using `[]` with `lapply()` can achieve something close to it.  


Example:
```{r}
df <- tibble(
  a = 1,
  b = 2,
  c = "a",
  d = "b",
  e = 4
)

# First find numeric columns
num_cols <- sapply(df, is.numeric)
num_cols

# Then transform each column with lapply() and replace the 
# original values.
df[, num_cols]
df
```

`sapply()` is similar to `lapply()`, but the former always tries to simplify the result, hence the `s` in the name.   

- The authors don't recommend using this for programming, because the simplification can fail and can give unexpected results. 

- Base R provides a stricter version of `sapply()`, called `vapply()` -- short for vector apply.  

Example using `vapply()` instead of `sapply()`:
```{r}
vapply(df, is.numeric, logical(1))
```

Example using `tapply()`, which computes a single grouped summary:
```{r}
diamonds |> 
  group_by(cut) |> 
  summarize(price = mean(price))

tapply(diamonds$price, diamonds$cut, mean)
```


## `for` loops

Basic structure:
```r
for (element in vector) {
  # do something with element
}
```

The most straighforward use of `for` loops is to achieve the same effect as `walk()` -- call some function with a side-effect on each element of a list. 

Example: 
```r
# version using walk
paths |> walk(append_file)

# version using for-loop
for (paths in paths) {
  append_file(paths)
}
```

To iterate over the list of file paths, it is better to iterate over their indices, rather than over the elements. To effectively do this, the function `seq_along()` can be used. The final function becomes:
```r
for (i in seq_along(paths)) {
 files[[i]] <- readxl::read_excel(paths[[i]])
}
```

Finally, to combine the list of tibbles into a single tibble, the combination of `do.call()` and `rbind()` can be used:
```r
do.call() + rbind()
```

To save the results, rather than make a list and save the results as we go, it is better to build the dataframe piece by peice:
```r
out <- NULL
for (path in paths) {
  out <- rbind(out, readxl::read_excel(path))
}
```

## Plots

There are two main types of base plot (for data science?) are scatterplots and histograms. In base R, we can make these plots using `plot()` and `hist()`.

Example:
```{r}
hist(diamonds$carat)
plot(diamonds$carat, diamonds$price)
```








