---
title: "Notes on Chapter 13 -- Numbers"
author: "Norman Lim"
date: "2025-07-27"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites
```{r}
library(tidyverse)
library(nycflights13)
```

## Making numbers

- parsing strings into numbers using  the `readr` package:
```{r}
x <- c("1.2", "5.6", "1e6")

parse_double(x)
```

- parsing when the intended string contains non-numeric text that needs to be ignored:
```{r}
x <- c("$1,234", "USD 3,513", "59%")

parse_number(x)
```


## Counts

- Using `count` for quick exploration and checks:
```{r}
flights |> count(dest)
```

- Using `count` to see the most common values:
```{r}
flights |> count(dest, sort = TRUE)
```

- To see all the results, we can add ` |> View()` or ` |> print(n = Inf)`

- Doing `count()` "by hand" using `group_by()`, `summarize()`, and `n()`. (Useful if you want to compute for other summaries too.)
```{r}
flights |> 
  group_by(dest) |> 
  summarize(
    n = n(),
    delay = mean(arr_delay, na.rm = TRUE)
  )
```


- Counting distinct values using `n_distinct()`:
```{r}
flights |> 
  group_by(dest) |> 
  summarize(carriers= n_distinct(carrier)) |> 
  arrange(desc(carriers))
```

- Getting a weighted count by summing:
```{r}
flights |> 
  group_by(tailnum) |> 
  summarize(miles = sum(distance))
```


- Getting a weighted count using the `wt` argument to `count()`:
```{r}
flights |> count(tailnum, wt = distance)
```


- Counting the number of missing values by combining `sum()` and `is.na()`:
```{r}
flights |> 
  group_by(dest) |> 
  summarize(n_cancelled = sum(is.na(dep_time)))
```

### Exercises
1. How can you use `count()` to count the number of rows with a missing value for a given variable?

Ans. 
Combine `count()` with `is.na()`.
```{r}
flights |> 
  count(is.na(dep_time)) |> 
  rename(
    is_na = `is.na(dep_time)`,
    count = n
  ) |> 
  mutate(is_na = if_else(is_na == FALSE, "not missing", "missing"))
```

2. Expand the following calls to `count()` to instead use `group_by()` `summarize()`, and `arrange()`:

> 1. `flights |> count(dest, sort = TRUE`)
> 2. `flights |>  count(tailnum, wt = distance`

Ans.

for No. 1
```{r}
flights |> 
  group_by(dest) |> 
  summarize(
    n = n()
  ) |> 
  arrange(desc(n))
```

for No. 2
```{r}
flights |> 
  group_by(tailnum) |> 
  summarize(miles = sum(distance))
```


### Arithmetic and recycling rules
- When the vector lengths are mismatched, the elements of the shorter vector is *recycled* or repeated:
```{r}
x <- c(1, 2, 10, 20)
x / 5
x / c(5, 5, 5, 5)
```

- R throws a warning when the length of the longer vector is not a multiple of the length of the shorter vector:
```{r}
x * c(1, 2)
x * c(1, 2, 3)
```

- Recycling rules are also applied to logical comparisons.


### Minimum and Maximum

- The functions `pmin()` and `pmax()` return the smallest or largest value in each row:
```{r}
df <- tribble(
  ~x, ~y,
  1, 3,
  5, 2,
  7, NA
)

df |> 
  mutate(
    min = pmin(x, y, na.rm = TRUE),
    max = pmax(x, y, na.rm = TRUE)
  )
```


### Modular arithmetic
 
- The function `%/%` gives the integer part of the quotient and `%%` gives the remainder:
```{r}
1:10 %/% 3
1:10 %% 3
```

- Modular arithmetic is handy for unpacking time into `hour` and `minute`:
```{r}
flights |> 
  mutate(
    hour = sched_dep_time %/% 100,
    minute = sched_dep_time %% 100,
    .keep = "used"
  )
```


- From section 12, we learned to determine the proportion of TRUE values to FALSE values in a logical vector by using `mean()`.  
- Combining this with modular arithmetic, we can plot the proportion of cancelled flights against the hour:
```{r}
flights |> 
  group_by(hour = sched_dep_time %/% 100) |> 
  summarize(
    prop_cancelled = mean(is.na(dep_time)),
    n = n()
  ) |> 
  filter(hour > 1) |> 
  ggplot(aes(x = hour, y = prop_cancelled)) +
  geom_line(color = "grey50") +
  geom_point(aes(size = n))
```


This is a condensed version of the code below:
```{r}
flights |> 
  mutate(hour = sched_dep_time %/% 100) |>
  group_by(hour) |> 
  summarize(
    prop_cancelled = mean(is.na(dep_time)), 
    n = n()
  ) |> 
  filter(hour > 1) |> 
  ggplot(aes(x = hour, y = prop_cancelled)) +
  geom_line(color = "grey50") +
  geom_point(aes(size = n))
```


### Logarithms

- Useful for dealing with data that ranges across multiple orders of magnitude.  
- `log2()` and `log10()` are generally easier to interpret than logs of other bases.  
- In base 2, every one step "forward" corresponds to doubling on the original scale, and every step "backward" corresponds to halving -- think of the f-stops in photography.  


### Rounding

- Rounding to the nearest integer:
```{r}
round(123.456)
```

- Changing the precision of rounding using the `digits` argument:
```{r}
round(123.456, digits = 2)  # two digits to the right of decimal point
round(123.456, 1)           # one digit...
round(123.456, -1)          # nearest ten
round(123.456, -2)          # nearest hundred
```

- R uses a rounding scheme called "Banker's rounding" -- a number halfway between two integers will be rounded to the even integer.
```{r}
round(c(1.5, 2.5))
```


- Because of this rounding scheme, we need to use `floor()` or `ceiling()` if we want to specifically round up or round down:
```{r}
x <- 123.456

floor(x)
ceiling(x)
```

- Since `floor()` and `ceiling()` don't have a `digits` argument, here's a trick you can use to control the number of digits used in rounding:
```{r}
# Round down to the nearest two digits
floor(x / 0.01) * 0.01

# Round up to the nearest two digits
ceiling(x / 0.01) * 0.01
```


### Cutting numbers into ranges

- Using `cut()` to break up a numeric vector into discrete buckets or bins:
```{r}
x <- c(1, 2, 5, 10, 15, 20)

# With even spacing
cut(x, breaks = c(0, 5, 10, 15, 20))

# With uneven spacing
cut(x, breaks = c(0, 5, 10, 100))

# breeaks with custom labels
cut(x,
  breaks = c(0, 5, 10, 15, 20),
  labels = c("small", "medium", "large", "extra_large")
)
```

- Any values outside the range of `breaks` will become `NA`
```{r}
y <- c(NA, -10, 5, 10, 30)
cut(y, breaks = c(0, 5, 10, 15, 20))

```


### Cummulative and rolling aggregates

- Base R provides functions that start with `cum`:
```{r}
x <- 1:10
cumsum(x)

s <- 1:100
cumsum(s)
```

This will make Gauss smile for sure.

- More functions for rolling or sliding aggregates are available in the `slider` package.


### Exercises
3. Currently `dep_time` and `sched_time` are convenient to look at, but hard to compute with because they're not really continuous numbers. You can see the basic problem by running the code below: there's a gap between each hour.
```{r}
flights |> 
  filter(month == 1, day == 1) |> 
  ggplot(aes(x = sched_dep_time, y = dep_delay)) +
  geom_point()
```

Convert them to a more truthful representation of time (either fractional hours or minutes since midnight).

Ans.
```{r}
flights |> 
  mutate(
    hour = sched_dep_time %/% 100,
    minute = sched_dep_time %% 100,
    decimal_time = hour + minute / 60
  ) |> 
  filter(month == 1, day == 1) |> 
  ggplot(aes(x = decimal_time, y = dep_delay)) +
  geom_point()
```

4. Round `dep_time` and `arr_time` to the nearest five minutes.

Ans.
```{r}
flights |> 
  mutate(
    dep_hour = dep_time %/% 100,
    dep_minute = sched_dep_time %% 100,
    dep_minute = round(dep_minute / 5) * 5,
    dep_hour = if_else(dep_minute == 60, dep_hour + 1, dep_hour),
    dep_minute = if_else(dep_minute == 60, 0, dep_minute),
    .after = dep_time
  )

# Similar approach for arr_time.
```


## General transformations

### Ranks

- `min_rank()` uses the typical method for dealing with ties:
```{r}
x <- c(1, 2, 2, 3, 4, NA)
min_rank(x)
```

- Using `desc()` to rank from highest to lowest:
```{r}
min_rank(desc(x))
```

- Examples of other ranking functions:
```{r}
df <- tibble(x = x)
df |> 
  mutate(
    row_number = row_number(x),
    dense_rank = dense_rank(x),
    percent_rank = percent_rank(x),
    cume_dist = cume_dist(x)
  )
```

Important arguments for `rank()` functions:
- `ties.method`: for breaking ties
- `na.last = "keep:`: to keep `NA`s as `NA`


- `row_number()`, when used without any arguments will give the "current" row.  
- We can take advantage of this when we need to divide data into similarly sized groups:
```{r}
df <- tibble(id = 1:10)

df |> 
  mutate(
    row0 = row_number() - 1,
    three_groups = row0 %% 3,
    three_in_each_group = row0 %/% 3
  )
```


### Offsets
- Using `lead()` and `lag()`:
```{r}
x <- c(2, 5, 11, 11, 19, 35)

lag(x)
lead(x)
```

Clever uses for `lag()`:
```{r}
# Get the difference between current and previous value
x - lag(x)

# Determine if the current value changes
x == lag(x)
```


### Consecutive identifiers

- Useful when we want the items to be in a different category after crossing a certain threshold.  
- Example: A situation where a new session should becin after a gap of more than x minutes since last activity:
```{r}
events <- tibble(
  time = c(0, 1, 2, 3, 5, 10, 12, 15, 17, 19, 20, 27, 28, 30)
)

events <- events |> 
  mutate(
    diff = time - lag(time, default = first(time)),
    has_gap = diff >= 5
  )

events
```

- Example: Similar situation as before, but we want to get numeric results so that we can apply other functions:
```{r}
events |> mutate(
  group = cumsum(has_gap)
)
```

- Example: Create a new grouping every time one of its arguments changes using `consecutive_id()` and `slice_head()`:
```{r}
df <- tibble(
  x = c("a", "a", "a", "b", "c", "c", "d", "e", "a", "a", "b", "b"),
  y = c(1, 2, 3, 2, 4, 1, 3, 9, 4, 8, 10, 199)
)

df |> 
  group_by(id = consecutive_id(x)) |> 
  slice_head(n = 1)
```

## Exercises
1. Find the 10 most delayed flights using a ranking function.
2. Which plane (`tailnum`) has the worst on-time record?
3. What time of day should you fly if you want to avoid delays as much as possible?
4. What does the code do?
```{r}
flights |> 
  group_by(dest) |> 
  filter(row_number() < 4)
```

Ans.
It returns only the first three rows of each group created by `group_by(dest)`. Approximately equal to 105 groups x 3.

What does this code do?
```{r}
flights |> 
  group_by(dest) |> 
  filter(row_number(dep_delay) < 4)
```

Ans. 
It returns only the first three rows of `dep_delay` for each group created by `group_by(dest)`.

5. For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.

Ans.

Total minutes of delay for each destination:
```{r}
flights |> 
  filter(arr_delay > 0) |> 
  group_by(dest) |> 
  summarize(
    total_arr_delay = sum(arr_delay, na.rm = TRUE)
  )
```

Proportion of total delay per destination for each flight:
```{r}
flights |> 
  filter(arr_delay > 0) |> 
  group_by(tailnum, dest) |> 
  summarize(
    ave_prop_arr_delay = mean(arr_delay / sum(arr_delay)),
    .groups = "drop"
  )
```

6. Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using `lag()`, explore how the average flight delay for an hour is related to the average delay for the previous hour.
```{r}
flights |> 
  mutate(hour = dep_time %/% 100) |> 
  group_by(year, month, day, hour) |> 
  summarize(
    dep_delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  mutate(lag = dep_delay - lag(dep_delay))
```

7. Look at each destination. Can you find flights that are suspiciously fast (i.e. flights that represent a potential data entry error)? Compute the air time of a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?

Ans.
Based on speed
```{r}
flights |> 
  mutate(
    speed = distance / (air_time / 60)
  ) |> 
  select(carrier, flight, dest, air_time, distance, speed) |> 
  arrange(desc(speed))
```

Since the ave. speed of commercial airlines, is usually between 500 to 600 mph, flights that are suspiciously fast or suspiciously slow can be flagged.

We can also do this by getting the standard deviation of `air_time`. Flights whose air time exceed the standard deviation by a certain threshold can be flagged.

Data entry checking based on air time of flight relative to the shortest flight:
```{r}
# Showing the top n flights which took longer than the shortest air time
flights |>
  select(carrier, flight, dest, air_time, distance) |> 
  group_by(flight, dest) |> 
  mutate(
    shortest_air_time = if_else(air_time == min(air_time), air_time, min(air_time)),
    prop_to_shortest = air_time / shortest_air_time,
  ) |> 
  arrange(desc(prop_to_shortest)) |> 
  head(10)
```




8. Find all destinations that are flown by at least two carriers. Use those destinations to come up with a relative ranking of carriers based on their performance for the same destination.

Ans. 
```{r}
# ranking carriers based on least amount of delays
flights |> 
  select(carrier, dest, dep_delay) |> 
  group_by(dest, carrier) |> 
  mutate(
    n = n()
  ) |> 
  filter(n >= 2) |> 
  summarize(
    ave_dep_delay = mean(dep_delay, na.rm = TRUE)
  ) |> 
  mutate(
    ranking = rank(ave_dep_delay)
  ) |> 
  arrange(dest,ranking) |> 
  # show the top 4 per destination
  filter(row_number() < 5)
```

## Numeric summaries

### Measures of central tendencies
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    mean = mean(dep_delay, na.rm = TRUE),
    median = median(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  ggplot(aes(x = mean, y = median)) +
  geom_abline(slope = 1, intercept = 0, color = "white") +
  geom_point()
```

- There is no **mode** function in R; this is not useful in real-wold data and statisticians tend not to use this function.  

### Minimum, maximum, and quantiles

Looking at the the 95% quantile of a data feature while ignoring the extreme 5% of cases:
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    max = max(dep_delay, na.rm = TRUE),
    q95 = quantile(dep_delay, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

### Spread

- `IQR()` is basically `quantile(x, 0.75) - quantile(x, 0.25).  

Exploring the curious case of EGE Airport using `IQR`:
```{r}
flights |> 
  group_by(origin, dest) |> 
  summarize(
    distance_iqr = IQR(distance),
    n = n(),
    .groups = "drop"
  ) |> 
  filter(distance_iqr > 0)
```


### Distributions

- Frequency plots of `dep_delay` for each day, for 365 days, overlaid, used to check if the distributions for subgroups resemble the whole:
```{r}
flights |> 
  filter(dep_delay < 120) |> 
  ggplot(aes(x = dep_delay, group = interaction(day, month))) +
  geom_freqpoly(binwidth = 5, alpha = 1/5)
```


- Whenever creating numerical summaries, it is a good idea to include the number of observations in each group.  

### Extracting values at nth position
 
- Example for extracting the `nth` value using position functions `first`, `last`, and `nth`:
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    first_dep = first(dep_time, na_rm = TRUE),
    fifth_dep = nth(dep_time, 5, na_rm = TRUE),
    last_dep = last(dep_time, na_rm = TRUE)
  )
```

- Be careful when using `first`, `nth`, and `last` -- the argument for removing `NA`s is `na_rm` and *not* `na.rm`.

- Extracting values at positions using filtering on ranks:
```{r}
flights |> 
  group_by(year, month, day) |> 
  mutate(r = min_rank(sched_dep_time)) |> 
  filter(r %in% c(1, max(r)))
```


### Summary functions using `mutate()`

- Useful when a standardized summary function is preferred.

Examples of `mutate` commands for summarizing data:  
- `x / sum(x)` for getting the proportion of total.  
- `(x - mean(x)) / sd(x)` for calculating the z-score (standardized to have a mean of 0 and sd of 1).  
- `(x - min(x)) / (max(x) - min(x))` for standardizing the data range to [0, 1].  
- `x / first(x)` for generating an index based on the first observation.



