---
title: "Notes on Chapter 25 - Functions"
author: "Norman Lim"
date: "2025-08-15"
output:
  pdf_document: default
  html_document: default
---


### Prerequisites
```{r}
library(tidyverse)
library(nycflights13)
```


## Vector functions

Vector functions are functions that take one or more vectors and return a vector result.  

Example:  
```{r}
df <- tibble(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5),
)

df |> mutate(
  a = (a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE)),
  b = (b - min(a, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE)),
  c = (c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE)),
  d = (d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE))
)
```


There is a mistake in `b`. This could have been avoided if we used functions. 

Three things needed to make a function:  
- A **name**.  
- The **arguments**.  
- The **body**.  

Format:  
```r
name <- function(x) {
  body
}
```

Creating the function for the rescaling code above:
```{r}
rescale01 <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
```


Testing if the function works as intended:
```{r}
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

Using the function (or calling the function):
```{r}
df |> mutate(
  a = rescale01(a),
  b = rescale01(b),
  c = rescale01(c),
  d = rescale01(d)
)
```


### Improving the function

Replacing the repetitious parts of the function:
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

Retesting:
```{r}
x <- c(1:10,Inf)
rescale01(x)
```

The result is not particularly useful. It is better to ignore infinite values:
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

Retesting:
```{r}
rescale01(x)
```


### Mutate functions

A simple variation of `rescale01`:
```{r}
z_score <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}
```

Another variation using `case_when()`:
```{r}
clamp <- function(x, min, max) {
  case_when(
    x < min ~ min,
    x > max ~ max,
    .default = x
  )
}

clamp(1:10, min = 3, max = 7)
```


A function that makes the first character upper case:
```{r}
first_upper <- function(x) {
  str_sub(x, 1, 1) <- str_to_upper(str_sub(x, 1, 1))
  x
}

first_upper("hello")
```

A function for stripping percent signs:
```{r}
# https://twitter.com/NVlabormarket/status/1571939851922198530
clean_number <- function(x) {
  is_pct <- str_detect(x, "%")
  num <- x |> 
    str_remove_all("%") |> 
    str_remove_all(",") |> 
    str_remove_all(fixed("$")) |> 
    as.numeric()
  if_else(is_pct, num / 100, num)
}

clean_number("$12,300")
clean_number("45%")
```

A function for replacing 997, 998, or 999 with `NA`:
```{r}
fix_na <- function(x) {
  if_else(x %in% c(997, 998, 999), NA, x)
}
```


### Summary functions

These are functions that return a single value for use in `summarize()`.

Example:
```{r}
commas <- function(x) {
  str_flatten(x, collapse = ", ", last = " and ")
}

commas(c("cat", "dog", "pigeon"))
```

Another example with mathematical computation -- coefficient of variation:
```{r}
cv <- function(x, na.rm = FALSE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}

cv(runif(100, min = 0, max = 50))
cv(runif(100, min = 0, max = 500))
```


An example for making handling missing data more "readable":
```{r}
n_missing <- function(x) {
  sum(is.na(x))
}
```

Another example with multiple vector inputs:
```{r}
mape <- function(actual, predicted) {
  sum(abs((actual - predicted) / actual)) / length(actual)
}
```


## Data frame functions

Embracing (wrapping a variable in double braces) to tell dplyr to use the value stored inside the argument:
```{r}
grouped_mean <- function(df, group_var, mean_var) {
  df |> 
    group_by({{ group_var }}) |> 
    summarize(mean({{ mean_var }}))
}

df <- tibble(
  mean_var = 1,
  group_var = "g",
  group = 1,
  x = 10,
  y = 100
)

df |> grouped_mean(group, x)
```

Without embracing, the function above will throw an error. This problem happens because dplyr uses **tidy evaluation** to allow referring to the names of variables inside the data frame without any special treatment.


### When to use embracing?

- **Data-masking:** used in functions like `arrange()`, `filter()`, and `summarize()` that compute with variables.  
- **Tidy-selection:** used for functions like `select()`, `relocate()`, and `rename()` that select variables.  


### Common use cases of embracing

```{r}
summary6 <- function(data, var) {
  data |> summarize(
    min = min({{ var }}, na.rm = TRUE),
    mean = mean({{ var }}, na.rm = TRUE),
    median = median({{ var }}, na.rm = TRUE),
    max = max({{ var }}, na.rm = TRUE),
    n = n(),
    n_miss = sum(is.na({{ var }})),
    .groups = "drop"
  )
}

diamonds |> summary6(carat)
```

**Tip:** Whenever you wrap `summarize()` in a helper, it is a good idea to set `.groups = "drop"` to both avoid seeing the warning and leaving the data in an ungrouped state.

Example using the function above on grouped data:
```{r}
diamonds |> 
  group_by(cut) |> 
  summary6(carat)
```

Example using the function above on computed variables:
```{r}
diamonds |> 
  group_by(cut) |> 
  summary6(log10(carat))
```


Example using the helper function `count()` to compute proportions:
```{r}
count_prop <- function(df, var, sort = FALSE) {
  df |> 
    count({{ var }}, sort = sort) |> 
    mutate(prop = n / sum(n))
}

diamonds |> count_prop(clarity)
```


Example with conditional variables as input argument:
```{r}
unique_where <- function(df, condition, var) {
  df |> 
    filter({{ condition }}) |> 
    distinct({{ var }}) |> 
    arrange({{ var}})
}

# Find all the destinations in December
flights |> unique_where(month == 12, dest)
```

Example of a function that automatically selects a predefined set of columns (aka features):
```{r}
subset_flights <- function(rows, cols) {
  flights |> 
    filter({{ rows }}) |> 
    select(time_hour, carrier, flight, {{ cols }})
}
```


### Data-masking vs. tidy-selection

Using the `pick()` function to make tidy-selection work inside a data-masking function: 
```{r}
count_missing <- function(df, group_vars, x_var) {
  df |> 
    group_by(pick({{ group_vars }})) |> 
    summarize(
      n_miss = sum(is.na({{ x_var}})),
      .groups = "drop"
    )
}

flights |> 
  count_missing(c(year, month, day), dep_time)
```

Example using `pick()` to make 2d table of columns:
```{r}
count_wide <- function(data, rows, cols) {
  data |> 
    count(pick({{ rows }}, {{ cols }})) |> 
    pivot_wider(
      names_from = {{ cols }},
      values_from = n,
      names_sort = TRUE,
      values_fill = 0
    )
}

diamonds |> count_wide(c(clarity, color), cut)
```


### Exercises

1. Using the datasets from `nycflights13, write a function that:

> a. Finds all flights that were cancelled.  
> b. Counts the number of cancelled flights and the number of flights delayed by more than an hour.  
> c. Finds all flights that were cancelled or delayed by more than a user-supplied number of hours.  
> d. Summarizes the weather to compute the minimum, mean, and maximum, of a user-supplied variable:  
> e. Converts the user-supplied variable that uses clock time (e.g. `dep_time`, `arr_time`, etc.) into a decimal time (i.e. `hours` + (`minutes`/60)).  

2. For each of the following functions, list all arguments that use tidy evaluation and describe whether they use data-masking or tidy-selection: `distinct()`, `count()`, `group_by()`, `rename_width()`, `slice_min()`, `slice_sample()`.  

3. Generalize the following function so that you can supply any number of variables to count.  
```r
count_prop <- function(df, var, sort = FALSE) {
  df |> 
    count({{ var }}, sort = sort) |> 
    mutate(prop = n / sum(n))
}
```

## Plot functions 

Instead of returning a value or a data frame, these functions return a plot. This works because `aes()` is a data-masking function. 

Example: Imagine if you need to make a lot histograms like this:
```{r}
diamonds |> 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

diamonds |> 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.05)
```

Using a function to generate the plots above (embracing is used because `aes()` is a data-masking function!):
```{r}
histogram <- function(df, var, binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth)
}

diamonds |> histogram(carat, 0.1)
```

The function above generates a `ggplot2` plot so additional components can be layered, just like in other plots:
```{r}
diamonds |> 
  histogram(carat, 0.1) +
  labs(
    x = "Size (in carats)",
    y = "Number of diamonds"
  )
```


### More variables

Example with overlaid plots:
```{r}
linearity_check <- function(df, x, y) {
  df |> 
    ggplot(aes(x = {{ x }}, y = {{ y }})) +
    geom_point() +
    geom_smooth(
      method = "loess",
      formula = y ~ x,
      color = "red",
      se = FALSE
    ) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      color = "blue", 
      se = FALSE
    )
}

starwars |> 
  filter(mass < 1000) |> 
  linearity_check(mass, height)
```


Example using a function to generate a plot for a large dataset where overplotting is a problem: 
```{r}
hex_plot <- function(df, x, y, z, bins = 20, fun = "mean") {
  df |> 
    ggplot(aes(x = {{ x }}, y = {{ y }}, z = {{ z }})) +
    stat_summary_hex(
      aes(color = after_scale(fill)), # make border color same as fill
      bins = bins,
      fun = fun
    )
}

diamonds |>  hex_plot(carat, price, depth)
```


### Combining with other tidyverse

A function to plot a sorted bar chart:
```{r}
sorted_bars <- function(df, var) {
  df |> 
    mutate({{ var }} := fct_rev(fct_infreq({{ var }}))) |> 
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}

diamonds |> sorted_bars(clarity)
```

Note: the operator `:=`, commonly referred to as the "walrus operator" was used in the example to generate the variable name based on user-supplied data. This is necessary because the common behavior in R is to have only literal names on the left of the `=` operator. The walrus operator is a workaround solution to this problem.  

Another example using functions to make it easier to draw a bar plot just for a subset of the data:
```{r}
conditional_bars <- function(df, condition, var) {
  df |> 
    filter({{ condition }}) |> 
    ggplot(aes(x = {{ var }})) +
    geom_bar()
}

diamonds |> conditional_bars(cut == "Good", clarity)
```


### Labeling

Using `rlang::englue()` to customize plot labels -- this function works in a similar way to how Python's f-strings are used in plot labels.  
```{r}
histogram <- function(df, var, binwidth) {
  label <- rlang::englue("A histogram of {{var}} with binwidth {{binwidth}}")
  df |> 
    ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth) +
    labs(title = label)
}

diamonds |> histogram(carat, 0.1)
```


### Exercises

Build up a rich plotting function by incrementally implementing each of the steps below:  

1. Draw a scatterplot given dataset and `x` and `y` variables.  
2. Add a line of best fit (i.e. a linear modwel with no standard errors).  
3. Add a title.  


## Style

General guidelines:  
- The name of the function should clearly describe what the function does.  
- Function names are generally verbs (except when they compute some very well-known nouns like `mean()`, or property of an object, etc.).  
- Argument names are generally nouns.  
- Put one space before the brackets.  
- Indent the body by two spaces.  
- Put an extra space before and after the variable name inside the embracing brackets (`{{}}`). 







