---
title: "Notes on Ch1 - Data Visualization"
author: "N_Lim"
date: "2025-06-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

## First Steps

```{r}
penguins
```

Dataset has 8 columns. We can also inspect the data using `glimpse()`:
```{r}
glimpse(penguins)
```

### Creating a ggplot
- define plot object using `ggplot()`
- then add layers to it

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

- ggplot2 follows the R philosophy that missing values should never go silently missing 


Adding colors
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  )
) +
  geom_point()
```

Adding a smooth curve
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  ) 
) +
  geom_point() +
  geom_smooth(method = "lm")
```


Using different shapes for each species:
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g
  )
) +
  geom_point(mapping = aes(
    color = species,
    shape = species
  )) +
  geom_smooth(method = "lm")
```

Improving the labels:
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g
  )
) +
  geom_point(mapping = aes(
    color = species,
    shape = species
  )) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    shape = "Species"
  )
```


Adding a theme to the plot:
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g
  )
) +
  geom_point(mapping = aes(
    color = species,
    shape = species
  )) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensins of Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    shape = "Species"
  ) +
  scale_color_colorblind()
```

Practice problem
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = flipper_length_mm,
    y = body_mass_g
  )
) +
  geom_point(mapping = aes(
    color = bill_depth_mm
  )) +
  geom_smooth()
```

## Visualizing distributions

Basic bar plot for categorical variable:
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```

Bar plot with the bars reordered according to count or frequency:
```{r}
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```

Basic histogram plot for visualizing the distribution of a variable
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

Basic density plot aka a smoothed-out version of the histogram (a histogram with infinite number of bars):
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

Basic boxplot
```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

Muliple density plots
```{r}
ggplot(penguins, aes(
  x = body_mass_g,
  color = species
)) +
  geom_density(linewidth = 0.75)
```

Multiple density plots with color fill:
```{r}
ggplot(penguins, aes(
  x = body_mass_g,
  color = species,
  fill = species
)) +
  geom_density(alpha = 0.5)
```

Basic stacked bar plots
```{r}
ggplot(penguins, aes(
  x = island,
  fill = species
)) +
  geom_bar()
```

Basic relative frequency plot aka stacked bar plot with the y-axis showing the proportion:
```{r}
ggplot(penguins, aes(
  x = island,
  fill = species
)) +
  geom_bar(position = "fill")
```

- easier to do in R than in Python
- the "count" label on the y-axis is now misleading and should be changed using `labs()`


Plotting using three or more variables
```{r}
ggplot(
  penguins,
  aes(
    x = flipper_length_mm,
    y = body_mass_g
  )
) +
  geom_point(aes(
    color = species,
    shape =island
  ))
```

- this is hard to look at, now that we have three variables
- a better way is to split the plots into facets

Splitting the plots into facets:
```{r}
ggplot(penguins, aes(
  x = flipper_length_mm,
  y = body_mass_g
)) +
  geom_point(aes(
    color = species,
    shape = island
  )) +
  facet_wrap(~island)
```

Saving plots
- use `ggsave()` 
```{r}
ggplot(penguins, aes(
  x = flipper_length_mm,
  y = body_mass_g
)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```

Notes on using `ggsave()`
- if the width and height are not specified, they will be taken from the dimensions of the current plotting device.
- the authors recommend using Quarto to assemble your final reports














