---
title: "Notes on Chapter 15 - Regular Expressions"
author: "Norman Lim"
date: "2025-07-29"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites
```{r}
library(tidyverse)
library(babynames)
```

## Pattern basics

- Using `str_view()` for basic string pattern matching:
```{r}
str_view(fruit, "berry")
```

- **Literal characters:** Letters and numbers that match exactly.  
- **Metacharacters:** Puntuation characters that have special meanings for pattern matching.  
- The dot (`.`) metacharacter will match any string that contains a character that come before it:
```{r}
str_view(c("a", "ab", "ae", "bd", "ea", "eab"), "a.")
```

Example: Find all the the words that contain an `a` followed by three letters, followed by an `e` in `fruits`:
```{r}
str_view(fruit, "a...e")
```

**Quantifiers:** control how many times a pattern can match:  
- `?`: makes a pattern optional (0 or once).
- `+`: lets a pattern repeat (at least once).
- `*`: lets a pattern be optional or repeat (any number of times, including 0).  

Examples:
```{r}
# matches an "a" optionally followed by a "b"
str_view(c("a", "ab", "abb"), "ab?")

# matches an "a" followed by at least one "b"
str_view(c("a", "ab", "abb"), "ab_")

# matches an "a" followed by any number of "b"s
str_view(c("a", "ab", "abb"), "ab*")
```

- **Character classes:** defined by `[]` and match a set of characters. Starting with `^` matches everyting except the character class -- the caret symbol acts like a "not" logic gate.
```{r}
str_view(words, "[aeiou]x[aeiou]")

str_view(words, "[^aeiou]l[^aeiou]")
```

- Using the **alternation** symbol `|` to pick one or more alternative patterns:
```{r}
str_view(fruit, "apple|melon|nut")

str_view(fruit, "aa|ee|ii|oo|uu")
```


## Key functions

### Detecting matches 

- Detecting matches using `str_detect()`:
```{r}
str_detect(c("a", "b", "c"), "[aeiou]")
```

- Using `str_detect()` in dplyr:
```{r}
babynames |> 
  filter(str_detect(name, "x")) |> 
  count(name, wt = n, sort = TRUE)
```

- Another example using `str_detect()` in dplyr (paired with `summarize()` this time):
```{r}
babynames |> 
  group_by(year) |> 
  summarize(prop_x = mean(str_detect(name, "x"))) |> 
  ggplot(aes(x = year, y = prop_x)) +
  geom_line()
```

### Counting matches

- Using `str_count()` to determine how many matches:
```{r}
x <- c("apple", "banana", "pear")

str_count(x, "p")
```

- Regex matches never overlap -- each match starts at the end of the previous match:
```{r}
str_count("abababa", "aba")
```

- Using `str_count()` in dplyr, paired with `mutate()`:
```{r}
babynames |> 
  count(name) |> 
  mutate(
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]")
  )
```

- The above result is not quite correct, since the default pattern matching is case-sensitive.  
Solution:
```{r}
babynames |> 
  count(name) |> 
  mutate(
    vowels = str_count(name, "[aeiouAIEOU]"),
    consonants = str_count(name, "[^aeiouAEIOU]")
  )
```

- Other applicable solutions:
> Use `vowels = str_count(name, regex("[aeiou]", ignore_case = TRUE))`
> Use `vowels = str_count(str_to_lower(name), "[aeiou]")

```{r}
babynames |> 
  count(name) |> 
  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]")
  )
```

### Replacing values

- Using `str_replace()` and/or `str_replace_all()`:
```{r}
x <- c("apple", "pear", "banana")

str_replace_all(x, "[aeiou]", "-")
```

- Using `str_remove()` and/or `str_remove_all()`:
```{r}
x <- c("apple", "pear", "banana")

str_remove_all(x, "[aeiou]")
```

### Extracting variables

- Using `separate_wider_` family of functions:
```{r}
df <- tribble(
  ~str,
  "<Sheryl>-F_34",
  "<Kisha>-F_45", 
  "<Brandon>-N_33",
  "<Sharon>-F_38", 
  "<Penny>-F_58",
  "<Justin>-M_41", 
  "<Patricia>-F_84", 
)

df |> 
  separate_wider_regex(
    str,
    patterns = c(
      "<",
      name = "[A-Za-z]+",
      ">-",
      gender = ".",
      "_",
      age = "[0-9]+"
    )
  )
```

### Exercises
1. What baby name has the most vowels? What name has the highest proportion of vowels? (Hint: what is the denominator?)

Ans.

baby name with the most vowels:
```{r}
babynames |> 
  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]")
  ) |> 
  filter(
    vowels == max(vowels)) |> 
  distinct(name, .keep = vowels) |> 
  rename(vowels = .keep)
  
  
```


Baby name with the highest proportion of vowels:
```{r}
babynames |> 
  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]"),
    length = str_length(name),
    v_prop = vowels / length
  ) |> 
  arrange(desc(v_prop)) |> 
  distinct(name, .keep_all = TRUE) |> 
  select(name, vowels, v_prop)
```


2. Replace all forward slashes in "a/b/c/d/e" with backslashes. What happens if you attempt to undo the transformation by replacing all backslashes with forward slashes? (We'll discuss the problem soon.)
```{r}
x2 <- "a/b/c/d/e"

str_replace(x2, "/", "\\")
```

This won't work as of now because slash is a special character. We need to use escaping keys.


3. Implement a simple version of `str_to_lower()` using `str_replace_all()`.
```{r}
x3 <- "thE qUIck And brOwn fOx"

x3 <- str_replace_all(x3, "[A]", "a")
x3 <- str_replace_all(x3, "[E]", "e")
x3 <- str_replace_all(x3, "[I]", "i")
x3 <- str_replace_all(x3, "[O]", "o")
x3 <- str_replace_all(x3, "[U]", "u")

str_view(x3)
```


4. Create a regular expression that will match telephone numbers as commonly written in your country.
```{r}
text = "lkjsdfljfds098234kljsd+63-927-5589flkjsdf09usdflkjsdflkjsdf"

pattern = "[+]63-[0123456789][0123456789][0123456789]-[0123456789][0123456789][0123456789][0123456789]"
str_view(text, pattern)
```


### Pattern details

### Escaping

- Escaping using backslash `\`:
```{r}
# to match a "." we need to use "\\."
dot <- "\\."
str_view(dot)

str_view(c("abc", "a.c"), "a\\.c")
```

- To match a literal backslash (`\`), we need to use four backslashes (`\\\\`):
```{r}
x <- "a\\b"
str_view(x)

str_view(x, "\\\\")
```

- Using raw strings in this case lets us avoid one layer of escaping:
```{r}
str_view(x, r"{\\}")
```

- To match a literal character that is also a metacharacter (`.`, `$`, etc.), it is more practical to use a character class (`[.]`, `[&]`, etc.)
```{r}
str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")

str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
```


### Anchors

- Used if you want to match at the start or end of the string. Default is to match any part of the string.
```{r}
# match at the start of the string
str_view(fruit, "^a")

# match at the end
str_view(fruit, "a$")
```

- To force a regex to match only the full string, anchor it with both `^` and `$`:
```{r}
str_view(fruit, "apple")
str_view(fruit, "^apple$")
```

- To match the boundaries between words, `\b` is used. This works for the start or end of a word:
```{r}
x <- c("summary(x)", "summarize(df", "rowsum(x)", "sum(x)")
str_view(x, "sum")

str_view(x, "\\bsum\\b")
```

- When used alone, anchors will produce a zero-width match:
```{r}
str_view("abc", c("S", "^", "\\b"))
```

- Replacing a standalone anchor:
```{r}
str_replace_all("abc", c("$", "^", "\\b"), "--")
```


### Character class shorthand

- `-` defines a range. Example: `[a-z]` and `[0-9]`.  
- `\` escapes special characters. Example: `[\^\-\]]` matches `^`, `-`, or `]`.

Example with code:
```{r}
x <- "abcd ABCD 12345 -!@#%."

str_view(x, "[abc]+")
str_view(x, "[a-z0-9]+")
str_view(x, "[^a-z0-9]+")

str_view("a-b-c", "[a-c]")
str_view("a-b-c", "[a\\-c]")
```

Other shorthands:
- `\d` matches any digit.  
- `\D` matches anything that isn't a digit.  
- `\s` matches any whitespace.  
- `\S` matches anything that isn't a whitespace.  
- `\w` matches any "word" character.  
- `\W` matches any "non-word" character.  

Examples:
```{r}
x <- "abcd ABCD 12345 -!@#$."

str_view(x, "\\d+")
str_view(x, "\\D+")
str_view(x, "\\s+")
str_view(x, "\\S+")
str_view(x, "\\w+")
str_view(x, "\\W+")
```

### Quantifiers

- Quantifiers control how many times a pattern matches.  
- `{n}` matches exactly n times.  
- `{n,}` matches at least n times.    
- `{n, m}` matches between n and m times.  


On the order of operation:
- Quantifiers have high precedence.  
- Alternation has low precedence.  


### Grouping and capturing

- Capturing groups allow the user to use sub-components of the match.  
- **back reference:**: `\1` refers to the match contained in the first parenthesis, `\2` in the second parenthesis, and so on.

- Example using back reference to find all fruits that have a repeated pair of letters:
```{r}
str_view(fruit, "(..)\\1")
```


- Example using back reference to find all words that start and end with the same pair of letters:
```{r}
str_view(words, "^(..).*\\1$")
```

- Using back references in `str_replace()`:
```{r}
sentences |> 
  str_replace("(\\w+) (\\w+) (\\w+)", "\\1 \\3 \\2") |> 
  str_view()
```

- Extracting matches for each group:
```{r}
sentences |> 
  str_match("the (\\w+) (\\w+)") |> 
  head()
```

- Converting the above result to a tibble:
```{r}
sentences |> 
  str_match("the (\\w+) (\\w+)") |> 
  as_tibble(.name_repair = "minimal") |> 
  set_names("match", "word1", "word2")
```


- Using `(?:)` to have parentheses without creating matching groups (aka non-capturing group):
```{r}
x <- c("a gray cat", "a grey dog")

# with capturing group
str_match(x, "gr(e|a)y")

# with non-capturing group
str_match(x, "gr(?:e|a)y")
```

### Exercises

...


## Pattern control

- Used when extra control over the details of the match is needed.  

### Regex flags

- `ignore_case = TRUE` flag is probably the most useful because it allows characters to match both uppercase or lowercase forms.  
```{r}
bananas <- c("banana", "Banana", "BANANA")

# no regex flag
str_view(bananas, "banana")

# with `ignore_case = TRUE` flag
str_view(bananas, regex("banana", ignore_case = TRUE))
```

- Flags for working with multiline strings:
- `dotall = TRUE` lets `.` match everything including `\n`.  
```{r}
x <- "Line 1\nLine 2\nLine 3"

# no flag
str_view(x, ".Line")

# with `dotall = TRUE` flag
str_view(x, regex(".Line", dotall = TRUE))
```

- `multiline = TRUE` makes `^` and `&` match the start and end of each line rather than those of the complete string: 
```{r}
x <- "Line 1\nLine 2\nLine 3"

# no flag
str_view(x, "^Line")

# with `multiline = TRUE` flag
str_view(x, regex("^Line", multiline = TRUE))
```

- `comments = TRUE` allows the use of comments and whitespace to make complex regex more understandable:
```{r}
phone <- regex(
  r"(
    \(?     # optional opening parens
    (\d{3}) # area code
    [)\-]?  # optional closing parens or dash
    \ ?     # optional space
    (\d{3}) # another three numbers
    [\ -]?  # optional space or dash
    (\d{4}) # four more numbers
  )", 
  comments = TRUE
)

str_extract(c("514-791-8141", "(123) 456 7890", "123456"), phone)
```

### Fixed matches

- `fixed()` is used to opt-out of the regex rules:
```{r}
str_view(c("", "a", "."), fixed("."))
```

- `fixed()` can also be used to ignore cases:
```{r}
str_view("x X", "X")

str_view("x X", fixed("X", ignore_case = TRUE))
```

- when working with non-English text, `coll()` works better than `fixed()`:
```{r}
str_view("i İ ı I", fixed("İ", ignore_case = TRUE))
str_view("i İ ı I", coll("İ", ignore_case = TRUE, locale = "tr"))

```


## Practice

### Check your work

Exercise: Finding all sentences that start with "The":
```{r}
str_view(sentences, "^The")
```

- We can see that this is not good enough. Words like `These` and `They` are also matches.
- Adding a word boundary:
```{r}
str_view(sentences, "^The\\b")
```

Looks OK!

- Exercise: Finding all sentences that begin with a pronoun:
```{r}
str_view(sentences, "^She|he|It|They\\b")
```

This is not good enough. 

- It is a good idea here to create a positive and negative matches and use them to test if your pattern works as expected:
```{r}
pos <- c("He is a boy.", "She had a good time")
neg <- c("Shells come from the sea", "Hadley said 'It's a great day'")

pattern <- "^(She|He|It|They)\\b"
str_detect(pos, pattern)
str_detect(neg, pattern)
```


### Boolean operations

Example: Find words that only contain consonants 
```{r}
str_view(words, "^[^aeiou]+$")
```

- This works but there is an easier way -- by flipping the problem around using the "not" operator.  
- Instead of looking for consonants, we can look for words that don't contain any vowels:
```{r}
str_view(words[!str_detect(words, "[aeiou]")])
```

Example: Find all words that contain "a" and "b".
- Plan of attack: look for all words that contain an "a" followed by a "b" or a "b" followed by an "a":
```{r}
str_view(words, "a.*b|b.*a")
```

- This plan works but is rather difficult to come up with. There is an easier way:
```{r}
words[str_detect(words, "a") & str_detect(words, "b")]
```

Example: Try to see if there was a word that contains all vowels.  
- Using the first plan of attack is not practical, since this requires 120 different patterns (5!).  
- a better way is to use the second approach:
```{r}
words[
  str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") &
  str_detect(words, "u")
]
```


- If you get stuck trying to create a single regex pattern that solves your problem, take a step back and think if you could break the problem down into smaller pieces.

### Creating a pattern with code

Problem: Find all sentences that mention a color.
```{r}
str_view(sentences, "\\b(red|green|blue)\\b")
```

What if we have more colors? Things would get cumbersome very quickly. 

- Better solution: store the colors in a vector and use `str_c` and `str_flatten()` to generate the necessary pattern.
```{r}
rgb <- c("red", "green", "blue")

str_c("\\b(", str_flatten(rgb, "|"), ")\\b")
```

- It works, but where can we get a list of colors?  
- There is a color list in R, but some items are numbered.
```{r}
str_view(colors())
```

- Eliminating the numbered colors:
```{r}
cols <- colors()
cols <- cols[!str_detect(cols, "\\d")]
str_view(cols)
```

Finally we can use this to generate the pattern:
```{r}
pattern <- str_c("\\b(", str_flatten(cols, "|"), ")\\b")
str_view(sentences, pattern)
```


### Exercises

...


## Regex in other places

### tidyverse

Some useful places where regex is useful
- `matches(pattern)` will select all available variables whose name matches the supplied pattern. Abvailable in `select()`, `rename_with()`, and `across()`.  
- `names_pattern` argument in `pivot_longer()`. Useful for extracting data out of variable names with complex structure.  
- `delim` argument in `separate_longer_delim()` and family of functions. 

### Base R

Some useful applications of regex in base R:
- `apropos(pattern)` searches all objects available in the global environment. Useful if you can't quite remember the name of a function.  
- `list.files(path, pattern)` lists all files in `path` that math a regex pattern.


