---
title: "Notes on Chapter 20 - Spreadsheets"
author: "The Caveman Coder"
date: "2025-08-12"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites
```{r}
library(readxl)
library(tidyverse)
library(writexl)
```

Reading Excel spreadsheets using the **readxl** package:  
- `read_xls()` reads Excel files with `xls` format.  
- `read_xlsx()` reads Excel files with `xlsx` format.  
- `read_excel()` reads both `xls` and `xlsx` formats. It infers the type based on the input.  
Examples:
```{r}
students <- read_excel("data/students.xlsx")
students
```


Reading an Excel file while supplying column names at the same time:
```{r}
read_excel(
  "data/students.xlsx",
  col_names = c("student_id", "full_name", "favorite_food", "meal_plan", "age")
)
```

Reading an Excel file while supplying column names at the same time, and skipping the original header:
```{r}
read_excel(
  "data/students.xlsx",
  col_names = c("student_id", "full_name", "favorite_food", "meal_plan", "age"),
  skip = 1
)
```


Reading an Excel file while supplying column names at the same time, skipping the original header, and specifying the values for `NA`:
```{r}
read_excel(
  "data/students.xlsx",
  col_names = c("student_id", "full_name", "favorite_food", "meal_plan", "age"),
  skip = 1,
  na = c("", "N/A")
)
```

Reading an Excel file while supplying column names at the same time, skipping the original header, and specifying the values for `NA`, and supplying the correct column types:
```{r}
students <- read_excel(
  "data/students.xlsx",
  col_names = c("student_id", "full_name", "favorite_food", "meal_plan", "age"),
  skip = 1,
  na = c("", "N/A"),
  col_types = c("numeric", "text", "text", "text", "text")
)

students <- students |> 
  mutate(
    age = if_else(age == "five", "5", age),
    age = parse_number(age)
  )

students
```

Note: we first read the `age` column as text, then corrected the issue, and then parsed it as numeric.


### Reading worksheets

Specifying which "tab" to open using the `sheet` argument:
```{r}
read_excel("data/penguins.xlsx", sheet = "Torgersen Island")
```

Repeting the operation above, with the `NA`s specified:
```{r}
penguins_torgersen <- read_excel(
  "data/penguins.xlsx", 
  sheet = "Torgersen Island",
  na = "NA"
)

penguins_torgersen
```

What if you don't know the worsheet names (aka the tab name)? Use `excel_sheets()`.
```{r}
excel_sheets("data/penguins.xlsx")
```

Reading the remaining worksheets:
```{r}
penguins_biscoe <- read_excel(
  "data/penguins.xlsx", 
  sheet = "Biscoe Island",
  na = "NA"
)
penguins_dream <- read_excel(
  "data/penguins.xlsx",
  sheet = "Dream Island", 
  na = "NA"
)
```

Checking the dimensions of each dataset extracted from the Excel worksheet:
```{r}
dim(penguins_torgersen)
dim(penguins_biscoe)
dim(penguins_dream)
```

Combining the files using `bind_rows()`:
```{r}
penguins <- bind_rows(
  penguins_torgersen,
  penguins_biscoe,
  penguins_dream
)

penguins
```


### Reading part of a sheet

Using `readxl_example()` to locate the spreadsheet in the directory:
`deaths_path <- readxl_example("death.xlsx")`

Reading excel while specifying the data range:
`read_excel(deaths_path, range = "A5:F15")`


### Writing to excel

Using `write_xlsx()`:
```{r}
bake_sale <- tibble(
  item = factor(c("brownie", "cupcake", "cookie")),
  quantity = c(10, 5, 8)
)

bake_sale

write_xlsx(bake_sale, path = "data/bake-sale.xlsx")
```

To create an Excel spreadsheet with more sophisticated formatting, we can use the **openxlsx** package. This package, however, is not part of tidyverse.


## Reading Google Sheets

### Prerequisites
```{r}
library(googlesheets4)
library(tidyverse)
```

We can use either the URL or the google sheet ID: `gs4_deauth()`

```{r}
gs4_deauth()
students_sheet_id <- "1V1nPp1tzOuutXFLb3G9Eyxi3qxeEhnOXUzL5_BcCQ0w"

students <- read_sheet(students_sheet_id)
students
```

Specifying the column names, `NA` stings, and column types while reading the spreadsheet:
```{r}
students <- read_sheet(
  students_sheet_id,
  col_names = c("student_id", "full_name", "favorite_food", "meal_plan", "age"),
  skip = 1,
  na = c("", "N/A"),
  col_types = "dcccc"
)

students
```

Example using the penguin dataset:
```{r}
penguins_sheet_id <- "1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY"
read_sheet(penguins_sheet_id, sheet = "Torgersen Island")
```

Worksheet names can be listed using `sheet_names()`:
```{r}
sheet_names(penguins_sheet_id)
```


### Writing to Google Sheets

Basic command:
```r
write_sheet(bake_sale, ss = "bake-sale")
```

Writing data to a specific worksheet inside a Google sheet:
```r
write_sheet(bake_sale, ss = "bake-sale", sheet = "Sales")
```

Reading from a public Google sheet without authentication with your Google account: `gs4_deauth()`

When you attempt to read in a sheet that requires authentication, googlesheets4 will direct you to a web browser with a prompt to sign in.  

To use an authentication token instead, the command is:
`gs4_auth(email = "xyz@google.com")`



