---
title: 'Notes on Chapter 12: Logical Vectors'
author: "Norman Lim"
date: "2025-07-24"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites
```{r}
library(tidyverse)
library(nycflights13)
```

Showing R's vectorized operation:
```{r}
x <- c(1, 2, 3, 5, 7, 11, 13)
x*2
```


- Any manipulation we can do to a free-floating vector, we can do to a variable inside a data frame using `mutate()`:
```{r}
df <- tibble(x)
df |> 
  mutate(y = x * 2)
```


## Comparisons

Using logical conditions:
```{r}
flights |> 
  filter(dep_time > 660 & dep_time < 2000 & abs(arr_delay) < 20)
```

Instad of doing it like this, `mutate()` provides a cleaner way to filter and create a new variable:
```{r}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
    .keep = "used"
  )
```


- This cleaner way makes it easier to both read the code and check if each step has been computed correctly (as intended).


Compare the code above with:
```{r}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20
  ) |> 
  filter(daytime & approx_ontime)
```


### Floating-point comparison

- Beware of using `==` when comparing numbers. Example:
```{r}
x <- c(1/49 * 49, sqrt(2)^2)
x
```

All seems OK until you test them for equality:
```{r}
x == c(1, 2)
```


Reason: The numbers contained in x are not exactly 1 and 2, respectively -- they are slightly off because there is no way to exactly represent `1/49` or `sqrt(2)`. We can see the exact values stored in `x` like this:
```{r}
print(x, digits = 16)
```


The `near` function can be used in this case -- it ignores small differences:
```{r}
near(x, c(1, 2))
```


### Missing values

- Missing values are "contagious" -- almost any operation involving an unknown will also be unknown:
```{r}
NA > 5
10 == NA
NA == NA
```


- For this reason, we cannot filter for `NA`s using `== NA`.

This will not work:
```{r}
flights |> 
  filter(dep_time == NA)
```

- This is the reason why we have the `is.na()` tool.


Using `is.na()`
```{r}
flights |> 
  filter(is.na(dep_time))
```

Sorting the results with `arrange()`:
```{r}
flights |> 
  filter(month == 1, day == 1) |> 
  arrange(dep_time)
```


- By default, `arrange()` puts all the `NA`s at the end. We can override this by arranging the results in descending order after filtering:
```{r}
flights |> 
  filter(month == 1, day == 1) |> 
  arrange(desc(is.na(dep_time)), dep_time)
```


### Exercises

1. How does `dplyr::near()` work? Type `near` to see the source code. Is `sqrt(2)^2` near 2?

Ans. `near` works by checking if the difference between the two values is less than the tolerance. If yes, then it is "near".

```{r}
near(sqrt(2)^2, 2)
```

2. Use `mutate()`, `is.na()`, and `count()` together to describe how the missing values in `dep_time`, `sched_dep_time` and `dep_delay` are connected.
```{r}
# test
```


## Boolean Algebra

- Don't use `&&` and `||` operators in `dplyr` functions. These are called short-circuiting operators. They are important for programming, not for data science.


### Order of operations

Instruction: "Find all flights that departed in November or December."

Wrong way of executing the instruction:
```{r}
flights |> 
  filter(month == 11 | 12)
```


Correct way of executing the instruction:
```{r}
flights |> 
  filter(month == 11 | month == 12)
```

- This mistake can easily happen because of the way we say it in our everyday language.


### `%in%`

- Instead of using `filter(x == y | x == z )` format (which is error-prone), we can simply use `%in%`.  
```{r}
flights |> 
  filter(month %in% c(11, 12))
```

- The `%in%` avoids the issue we had with `==` when used to check for `NA`s:
```{r}
# using ==
c(1, 2, NA) == NA
c(1, 2, NA) %in% NA
```

- The `%in%` expression can be used to filter for `NA`s:
```{r}
flights |> 
  filter(dep_time %in% c(NA, 0800))
```


## Exercises

1. Find all flights where `arr_delay` is missing but `dep_delay` is not. Find all flights where neither `arr_time` nor `sched_arr_time` are missing, but `arr_delay` is.

Ans.
```{r}
flights |> 
  filter(is.na(arr_delay) & !is.na(dep_delay))

flights |> 
  filter(!is.na(arr_time) & !is.na(arr_time) & is.na(arr_delay))
```


2. How many flights have a missing `dep_time`? What other variables are missing in these rows? What might these rows represent?

Ans.
```{r}
flights |> 
  filter(is.na(dep_time))

n_missing_dep_time <- flights |> 
  filter(is.na(dep_time)) |> 
  nrow()
```

There are `r n_missing_dep_time` flights with missing `dep_time`. `dep_delay` and `arr_time` are also missing, along with missing `dep_time`.


3. Assuming that a missing `dep_time` implies that a flight is cancelled, look at the number of cancelled flights per day. Is there a pattern? Is there a connection between the proportion of cancelled flights and the average delay of non-cancelled flights?


## Summaries

### Logical summaries

- Using `all()` and `any()`:
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    all_delayed = all(dep_delay <= 60, na.rm = TRUE),
    any_long_delay = any(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```


### Numerical summaries
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    proportion_delayed = mean(dep_delay <= 60, na.rm = TRUE),
    count_long_delay = sum(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```


### Logical subsetting

Condition: find the average delay just for the flights that were actually delayed.

Plan of attack: first, filter for the delayed flights. Then, calculate the average delay.
```{r}
flights |> 
  filter(arr_delay > 0) |> 
  group_by(year, month, day) |> 
  summarize(
    behind = mean(arr_delay),
    n = n(),
    .groups = "drop"
  )
```

Condition: In addition to the conditions above, also compute for the average delay for flights that arrived early.

Plan of attack: similar to the plan above, but perform a separate filter step. Then figure out how to combine the two dataframes altogether.

Better plan of attack: Perform an inline filter: `arr_delay[arr_delay > 0]` to get only the positive arrival delays. This is similar to how boolean filtering is usually performed in Python.
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    behind = mean(arr_delay[arr_delay > 0], na.rm = TRUE),
    ahead = mean(arr_delay[arr_delay < 0], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```


## Exercises
1. What will `sum(is.na(x))` tell you? How about `mean(is.na(x))`
2. What does `prod()` return when applied to a logical vector? What logical summary function is it equivalent to? What does `min()` return when returned to a logical vector? What logical summary is it equivalent to?

Ans.
```{r}
TT_vector <- c(TRUE, TRUE)
TF_vector <- c(TRUE, FALSE)
TFT_vector <- c(TRUE, FALSE, TRUE)
FTF_vector <- c(FALSE, TRUE, FALSE)

prod(TT_vector)
prod(TF_vector)
prod(TFT_vector)
prod(FTF_vector)

min(TT_vector)
min(TF_vector)
min(TFT_vector)
min(FTF_vector)

sum(all(TT_vector))
sum(all(TF_vector))
sum(all(TFT_vector))
sum(all(FTF_vector))
```

When applied to logical vectors, they act like `sum(all())`


### Conditional transformations

Example using `if_else()`
```{r}
x <- c(-3:3, NA)

if_else(x > 0, "+ve", "-ve")
```

The syntax is similar to the IF formula in Excel, but there is an optional argument specifies what to return instead of NA:
```{r}
if_else(x > 0, "+ve", "-ve", "???")
```


A simple version of the `coalesce()` function using `if_else()`
```{r}
x1 <- c(NA, 1, 2, NA)
y1 <- c(3, NA, 4, 6)

ifelse(is.na(x1), y1, x1)
```


- Just like in Excel, the `if_else()` function can be hard to read (and hard to code) when the conditions get complicated:
```{r}
if_else(x == 0, "0", if_else(x < 0, "+-ve", "+ve"), "???")
```

- For complicated conditions, it is better to use `case_when()` instead of `if_else()`:
```{r}
x <- c(-3:3, NA)

case_when(
  x == 0 ~ "0",
  x < 0 ~ "-ve",
  x > 0 ~ "+ve",
  is.na(x) ~ "???"
)
```

The code is longer but it more readable and easier to code.

- With `case_when()`, it is also possible to create a default/catch-all value to return:
```{r}
case_when(
  x < 0 ~ "-ve",
  x > 0 ~ "+ve",
  .default = "???"
)
```

Just like `vlookup` in Excel, if multiple conditions match, only the first case will be used:
```{r}
case_when(
  x > 0 ~ "+ve",
  x > 2 ~ "big"
)
```

- With `case_when()`, other logical conditions can be used:
```{r}
flights |> 
  mutate(
    status = case_when(
      is.na(arr_delay)     ~ "cancelled",
      arr_delay < -30      ~ "very early",
      arr_delay < -15      ~ "early",
      abs(arr_delay) <= 15 ~ "on time",
      arr_delay < 60       ~ "late",
      arr_delay < Inf      ~ "very late"
    ),
    .keep = "used"
  )
```


## Exercises
1. A number is even if it's divisible by two, which in R you can find out with x %% 2 == 0. Use this fact and `if_else()` to determine whether each number between 0 and 20 is even or odd.

Ans.
```{r}
x <- c(0:20)

if_else(x %% 2 == 0, "even", "odd")
```

2. Given a vector of days like `x <- c("Monday", "Saturday", "Wednesday"), use an `if_else()` statement to label them as weekends or weekdays.
```{r}
x <- c("Monday", "Saturday", "Wednesday", "March")
weekends <- c("Saturday", "Sunday")
days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

cond1 <- x %in% days
cond2 <- x %in% weekends
if_else(cond1, if_else(cond2, "weekends", "weekdays"), "???")
```

3. Use `if_else()` to compute the absolute value of a numeric vector called `x`.

Ans.
```{r}
x <- c(-3:3, NA)
if_else(x >= 0, x, -x)
```

4. Write a `case_when()` statement that uses the `month` and `day` columns from `flights` to label a selection of important US holidays (e.g., New Years Day, 4th of July, Thanksgiving, and Christmas). First create a logical column that is either `TRUE` or `FALSE`, and then create a character column that either gives the name of the holiday or is `NA`.

```{r}
flights |> 
  mutate(
    is_holiday = case_when(
      (day == 1 & month == 1) ~ TRUE,
      (day == 4 & month == 7) ~ TRUE,
      (day == 25 & month == 25) ~ TRUE,
      TRUE ~ NA
    ),
    holiday_type = case_when(
      (day == 1 & month == 1) ~ "New Years Day",
      (day == 4 & month == 7) ~ "Independence Day",
      (day == 25 & month == 12) ~ "Christmas Day"
    ),
    .keep = "used"
  )
```









