---
title: "Notes on Ch4: Workflow - code style"
author: "N. Lim"
date: "2025-06-30"
output:
  pdf_document: default
  html_document: default
---

- Good coding style is like correct punctuation; you can manage without it, but it sure makes things easier to read.  

- A good way to restyle code is to use the **styler** package.


```{r}
library(tidyverse)
library(nycflights13)
```


## R-rrific style for Names
- use lowercase letters, numbers, and `_`.
- use `_` to separate words  

Example:
```{r}
short_flights <- flights |> 
  filter(air_time < 60)
```


## R-rrific style for Spaces
- Put spaces on either side of mathematical operators execpt for `^`.

Example:
```r
z <- (a + b)^2 / d
```

- Don't put spaces inside or outside parentheses for regular function calls.
- Always put a space after a comma, just like in English.

Example:
```r
mean(x, na.rm = TRUE)
```

- It's ok to add extra spaces if it improves alignment.

Example:
```{r}
flights |> 
  mutate(
    speed      = distance / air_time,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %% 200
  )
```


## R-rrific style for Pipes
- Put spaces before and after the pipe (`|>` or `%>%`)

Example:
```{r}
flights |> 
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)
```

- If the function you're piping ino has named arguments, put each argument on a new line.  
- If the function doesn't have named arguments, keep everything on one line unless it doesn't fit, in which case you should put each argument on its own line.

Example:
```{r}
flights |> 
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```

- After the first step of the pipeline, indent each line by **two spaces**.
- If you're putting each argument on its own line, indent by an extra two spaces.

Example:
```{r}
flights |> 
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```

- It's ok to shirk some of these rules if your pipeline fits easily on one line. 

Examples:
```r
# This fits on one line
df |> mutate(y = x + 1)

# This is preferred because it is easier to add
# more variables in the future
df |> 
  mutate(
  y = x + 1
)
```

- Be wary of writing very long lines (more than 10-15 lines). It is better to break them up into smaller sub-tasks and give each task an informative name.


## R-rrific style for ggplot
- Treat the `+` the same way as the pipe.

Example:
```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = delay)) +
  geom_point() +
  geom_line()
```

- If you can't fit all of the arguments to a function on a single line, put each argument on its own line.

Example:
```{r}
flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(distance / air_time, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = distance, y = speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE,
    color = "white",
    linewidth = 4
  ) +
  geom_point()
```


## R-rrific style for sectioning comments
```{r}
# Load data --------------------------------------

# Plot data --------------------------------------
```

## Exercises

E1. Restyle the following pipelines following the guidelines above:

---r
flights|>filter(dest=="IAH")|>group_by(year,month,day)|>summarize(n=n(),
delay=mean(arr_delay,na.rm=TRUE))|>filter(n>10)

flights|>filter(carrier=="UA",dest%in%c("IAH","HOU"),sched_dep_time>
0900,sched_arr_time<2000)|>group_by(flight)|>summarize(delay=mean(
arr_delay,na.rm=TRUE),cancelled=sum(is.na(arr_delay)),n=n())|>filter(n>10)
---

Solution:
```{r}
flights |> 
  filter(dest == "IAH") |>
  group_by(year, month, day) |> 
  summarize(
    n = n(),
    delay = mean(arr_delay,na.rm=TRUE)
  ) |>
  filter(n > 10)
```
```{r}
flights |> 
  filter(
    carrier == "UA",
    dest %in% c("IAH", "HOU"),
    sched_dep_time > 0900,
    sched_arr_time > 2000
  ) |> 
  group_by(flight) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    cancelled = sum(is.na(arr_delay)),
    n = n()
  ) |> 
  filter(n > 10)
```

