---
title: "Notes on Chapter 21 - Databases"
author: "The Caveman Coder"
date: "2025-08-13"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites
```{r}
library(DBI)
library(dbplyr)
library(tidyverse)
```

Connecting to duckdb:
```{r}
con <- DBI::dbConnect(duckdb::duckdb())
```


### Loading data
```{r}
dbWriteTable(con, "mpg", ggplot2::mpg)
dbWriteTable(con, "diamonds", ggplot2::diamonds)
```

### DBI basics

Listing all tables in the database using `dbListTables()`:
```{r}
dbListTables(con)
```

Reading tables using `dbReadTable()`:
```{r}
con |> 
  dbReadTable("diamonds") |> 
  as_tibble()
```


Running SQL queries on the table using `dbGetQuery()`:
```{r}
sql <- "
  SELECT carat, cut, clarity, color, price
  FROM diamonds
  WHERE price > 15000
"

as_tibble(dbGetQuery(con, sql))
```


## dbplyr basics

Using `tbl()` to create a database object:
```{r}
diamonds_db <- tbl(con, "diamonds")
diamonds_db
```

The database object created uses *lazy evaluation*; it only performs the operation when needed.

For example,
```{r}
big_diamonds_db <- diamonds_db |> 
  filter(price > 15000) |> 
  select(carat:clarity, price)

big_diamonds_db
```


It doesn't know the number of rows! Why? Because knowing the total number of rows requires executing the complete query first.


Using the `show_query()` function to see the SQL code used to create the db object:
```{r}
big_diamonds_db |> 
  show_query()
```

Using `collect()` to get all the data back from a db object created using SQL query:
```{r}
big_diamonds <- big_diamonds_db |> 
  collect()

big_diamonds
```


## SQL

In this section, we will learn about the relationship between dplyr commands and SQL queries.

Copying the required data (using the built-in `copy_nycflights13()`) to our database:
```{r}
dbplyr::copy_nycflights13(con)

flights <- tbl(con, "flights")
planes <- tbl(con, "planes")
```

### SQL basics
```{r}
flights |> show_query()
planes |>  show_query()
```

Using `WHERE` and `ORDER BY` to control which rows are included and how they are ordered:
```{r}
flights |> 
  filter(dest == "IAH") |> 
  arrange(dep_delay) |> 
  show_query()
```

Showing how `GROUP BY` converts the query to a summary, causing aggregation to happen:
```{r}
flights |> 
  group_by(dest) |> 
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) |> 
  show_query()
```

Important distinction between dplyr verbs and SELECT clauses:  
- SQL is not case-sensitive. It is only by convention that we write queries in upper-case.  
- Order matters in SQL. The proper order is: `SELECT`, `FROM`, `WHERE`, `GROUP BY`, `ORDER BY`.  
- SQL evaluates the clauses in this order: `FROM`, `WHERE`, `GROUP BY`, `SELECT`, and `ORDER BY`.

### `SELECT`

- Performs the same job as `select()`, `mutate()`, `rename()`, `relaocate()`, and `summarize()`.

```{r}
planes |> 
  select(tailnum, type, manufacturer, model, year) |> 
  show_query()

planes |> 
  select(tailnum, type, manufacturer, model, year) |> 
  rename(year_built = year) |> 
  show_query()

planes |> 
  select(tailnum, type, manufacturer, model, year) |> 
  relocate(manufacturer, .before = type) |> 
  show_query()
```

In the query generated, "year" and "type" are wrapped in double quotes because these are **reserved words** in duckdb. Wrapping them in `"` avoids any potential confusion between column/table names and SQL operators.  

To be safe, we can wrap every column names in `"`. Other database systems use backticks instead of quotation marks.

- Example of `mutate()` translated into `SELECT` in SQL:
```{r}
flights |> 
  mutate(
    speed = distance / (air_time / 60)
  ) |> 
  show_query()
```

### `FROM`

This is more useful when we are joining tables.

### `GROUP BY`

The **dplyr** `group_by()` is translated to `GROUP BY` and `summarize()` is translated to `SELECT`:
```{r}
diamonds_db |> 
  group_by(cut) |> 
  summarize(
    n = n(),
    avg_price = mean(price, na.rm = TRUE)
  ) |> 
  show_query()
```


### `WHERE`

The **dplyr** function `filter()` is translated to `WHERE`:
```{r}
flights |> 
  filter(dest == "IAH" | dest == "HOU") |> 
  show_query()

flights |> 
  filter(arr_delay > 0 & arr_delay < 20) |> 
  show_query()
```

Notable differences:  
- the OR pipe `|` becomes `OR`.  
- the `&` symbol becomes `AND`.  
- comparison is performed using `=` instead of `==`.  
- for literal strings, single quotes (`''`) are used instead of double quotes (`"`) because double quotes are used to avoid conflicts with reserved words.  
- SQL uses `NULL` instead of `NA`.  
- the `%in%` operator somewhat becomes `IN` in SQL

Examples:
```{r}
flights |> 
  filter(!is.na(dep_delay)) |> 
  show_query()

flights |> 
  filter(dest %in% c("IAH", "HOU")) |> 
  show_query()
```

While the expression `WHERE (NOT((dep_delay IS NULL)))` works, it is not very "human-readable" -- a better way to express this is:
`WHERE "dep_delay" IS NOT NULL`.

In **dplyr**, we can use `filter()` after `summarize()`. This is translated to SQL queries by means of a `HAVING` clause:
```{r}
diamonds_db |> 
  group_by(cut) |> 
  summarize(n = n()) |> 
  filter(n > 100) |> 
  show_query()
```


### `ORDER BY`

Ordering rows is handled by `ORDER BY` in SQL:
```{r}
flights |> 
  arrange(year, month, day, desc(dep_delay)) |> 
  show_query()
```

The `desc()` funtion in **dplyr** is directly inspired by SQL's `DESC`.


### `Subqueries`

In SQL, a subquery is simply a query used as a data source in the `FROM` clause instead of the usual table.
```{r}
flights |> 
  mutate(
    year1 = year + 1,
    year2 = year1 + 1
  ) |> 
  show_query()

flights |> 
  mutate(year1 = year + 1) |> 
  filter(year1 == 2014) |> 
  show_query()
```

Note: sometimes, the translation to SQL is not yet optimized and plyr will create a subquery when it's not needed.


### `JOIN`

Joins in **dplyr** and SQL are similar:
```{r}
flights |> 
  left_join(planes |> rename(year_built = year), join_by(tailnum)) |> 
  show_query()
```


### Other verbs

To see the full set of what's available in dbplyr, go to https://dbplyr.tidiverse.org/reference/.


## Function translations

To help understand how column-wise functions get translated into SQL queries, the following helper functions are helpful:

```{r}
summarize_query <- function(df, ...) {
  df |> 
    summarize(...) |> 
    show_query()
}

mutate_query <- function(df, ...) {
  df |> 
    mutate(..., .keep = "none") |> 
    show_query()
}
```

Using the helper functions above to understand how complex operations in dplyr can be done using SQL queries:
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize_query(
    mean = mean(arr_delay, na.rm = TRUE),
    median = median(arr_delay, na.rm = TRUE)
  )
```

Here is another example using the function`OVER` in the translation:
```{r}
flights |> 
  group_by(year, month, day) |> 
  mutate_query(
    mean = mean(arr_delay, na.rm = TRUE)
  )
```

It is done this way in SQL because the `GROUP BY` clause in SQL is used exclusively for summaries --- the task of grouping is handed over to `OVER`.

The `OVER` clause can also "look" forward or backwards using `LEAD` and `LAG` clauses:
```{r}
flights |> 
  group_by(dest) |> 
  arrange(time_hour) |> 
  mutate_query(
    lead = lead(arr_delay),
    lag = lag(arr_delay)
  )
```

Since SQL tables are designed to handle large sets of data, the tables have no intrinsic order. In the **dplyr** code above, if the `arrange()` verb is not used, one might get different results each time the code is run.


The **dplyr** expressions `if_else()` and `case_when()` are translated to `CASE WHEN` in SQL:
```{r}
flights |> 
  mutate_query(
    description = if_else(arr_delay > 0, "delayed", "on-time")
  )

flights |> 
  mutate_query(
    description = case_when(
      arr_delay < -5 ~ "early",
      arr_delay < 5 ~ "on-time",
      arr_delay >= 5 ~ "late"
    )
  )
```

Other functions in R that don't have a direct translation to SQL can also be handled by `CASE WHEN`:
```{r}
flights |> 
  mutate_query(
    description = cut(
      arr_delay,
      breaks = c(-Inf, -5, 5, Inf),
      labels = c("early", "on-time", "late")
    )
  )
```


While **dbplyr** translations are not (yet) perfect, they work well especially on commonly-used functions.



