---
title: 'Notes on Chapter 14: Strings'
author: "Norman Lim"
date: "2025-07-28"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

### Prerequisites
```{r}
library(tidyverse)
library(babynames)
```

## Creating a string
```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```


### Escapes
```{r}
double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'"
```

- Using a literal backslash to escape a backlash:
```{r}
backslash <- "\\"
```

- To view strings properly, use `str_view()`:
```{r}
x <- c(single_quote, double_quote, backslash)

str_view(x)
```


### Raw strings

- Using multiple quotes and/or backslashes can be hard to read and confusing.  
- Better method is to declare the string as "raw" string, just like in Python:
```{r}
tricky <- r"(double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'")"

str_view(tricky)
```

- Raw strings in R mus be enclosed with `r"()"` or `r"[]"`` or `r"{}"`

### Other special characters

- `\n` for new line.  
- `\t` for tab.
- `\u` or `\U` for unicode escapes

```{r}
x <- c("one\ntwo", "one\ttwo", "\u00b5", "\U0001f604")

str_view(x)
```

- `str_view()` uses `{\t} for tabs to make them easier to spot.

### Exercises
1. Create strings that contain the following values:
> 1. He said "That's amazing!"
> 2. \a\b\c\d
> 3. \\\\\\

```{r}
# No. 1
str_1 <- r"("He said That's amazing!")"
str_view(str_1)

# No. 2
str_2 <- r"(\a\b\c\d)"
str_view(str_2)


# No. 3
str_3 <- r"(\\\\\\)"
str_view(str_3)

```

2. Create the string in your R session and print it. What happens to the special "\u00a0"? How does `str_view()` display it? Can you do a little googling to figure out what this special character is?

Ans.
```{r}
str_4 <- "This\u00a0is\u00a0tricky"
str_view(str_4)
```

The unicode character u+00a0 (`\u00a0`) stands for no-break space. This is a space character that prevents an automatic line break at its position. 


### Creating many strings from data

- `str_c`: takes any number of vectors as agruments and returns a character vector:
```{r}
str_c("x", "y")
str_c("x", "y", "z")
str_c("Hello, ", c("John", "Susan"))
```

- `str_c` is similar to `past0()`, but is designed to be used with `mutate()`:
```{r}
df <- tibble(name = c("Flora", "David", "Terra", NA))
df |> 
  mutate(greeting = str_c("Hi ", name, "!"))
```

- `coalesce()` displays `NA`s in a different way:
```{r}
df |> 
  mutate(
    greeting1 = str_c("Hi ", coalesce(name, "you"), "!"),
    greeting2 = coalesce(str_c("Hi", name, "!"))
  )
```

### `str_glue()`

- Using `str_c()` leads to a lot of `"`, making the code unreadable. `str_glue()` behaves in a similar way to f-stings in Python, but works in dplyr:
```{r}
df |> 
  mutate(greeting = str_glue("Hi {name}!"))
```

- One drawback of `str_glue()` is the way it handles `NA`s -- `str_glue()` converts them to the literal string "NA".  

- To escape the curly brackets character "{" or "}" in `str_glue()`, you double it:
```{r}
df |> 
  mutate(greeting = str_glue("{{Hi {name}!}}"))
```

### `str_flatten()`

- `str_flatten()` takes a character vector and combines each element of the vector into a single string:
```{r}
str_flatten(c("x", "y", "z"))
str_flatten(c("x", "y", "z"), ", ")
str_flatten(c("x", "y", "z"), ", ", last = ", and ")
```

- Using `str_flatten()` with `summarize()`:
```{r}
df <- tribble(
  ~ name, ~ fruit,
  "Carmen", "banana",
  "Carmen", "apple",
  "Marvin", "nectarine",
  "Terrence", "cantaloupe",
  "Terrence", "papaya",
  "Terrence", "mandarin"
)

df |> 
  group_by(name) |> 
  summarize(fruits = str_flatten(fruit, ", "))
```

### Exercises

1. Compare and contrast the results of `paste0` with `str_c()` for the following inputs:

```r
str_c("hi ", NA)
str_c(letters[1:2], letters[1:3])
```
```{r}
# str_c
str_c("hi ", NA)
# str_c(letters[1:2], letters[1:3])

# paste0
paste0("hi ", NA)
paste0(letters[1:2], letters[1:3])
```

`str_c()` cannot threw an error running the second line due to "incompatible size"


## Extracting data from strings

### Separating into rows

- The most common tool is `separate_longer_delim()`:
```{r}
df1 <- tibble(x = c("a,b,c", "d,e", "f"))
df1 |> 
  separate_longer_delim(x, delim = ",")
```

- Using `separate_longer_position()`:
```{r}
df2 <- tibble(x = c("1211", "131", "21"))
df2 |> 
  separate_longer_position(x, width = 1)
```

### Separating into columns

- Using `separate_wider_delim()`: 
```{r}
df3 <- tibble(x = c("a10.1.2022", "b10.2.2011", "e15.1.2015"))

df3 |> 
  separate_wider_delim(
    x, delim = ".",
    names = c("code", "edition", "year")
  )
```

- Using the argument `NA` in `searate_wider_delim()` to omit a piece from the source vector :
```{r}
df3 |> 
  separate_wider_delim(
    x, 
    delim = ".",
    names = c("code", NA, "year")
  )
```

- `separate_wider_position()` generally requires a known fixed set of columns:
```{r}
df4 <- tibble(x = c("202215TX", "202122LA", "202325CA"))
df4 |> 
  separate_wider_position(
    x,
    widths = c(year = 4, age = 2, state = 2)
  )
```

### Diagnosing widening problems

-If the width of a column is not known, the tools `too_few` and `too_many` can be useful:

```{r}
df <- tibble(x = c("1-1-1", "1-1-2", "1-3", "1-3-2", "1"))
```

Running the code below will not work:
```r
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z")
  )
```
This code will throw the following error:

```
#> Error in `separate_wider_delim()`:
#> ! Expected 3 pieces in each element of `x`.
#> ! 2 values were too short.
#> ℹ Use `too_few = "debug"` to diagnose the problem.
#> ℹ Use `too_few = "align_start"/"align_end"` to silence this message
```

- We can use these error messages to debug our code:
```{r}
debug <- df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_few = "debug"
  )
```

Looking at the contents of `debug`:
```{r}
debug
```

- The `x_ok` column tells us where the problem occurred:
```{r}
debug |> filter(!x_ok)
```

- The `x_pieces` column how many pieces were found. In this case we are expecting 3, but we only found 2 and 1 pieces. Now we know that the problem was due to having "too_few" pieces. We can fill them with `NA`s to avoid R from throwing an error.  
- We can then use the argument `too_few = "align_start"` or `too_few = "align_end"` to control where the `NA`s should go.  
```{r}
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_few = "align_start"
  )
```

What if we have too many pieces?
```{r}
df <- tibble(x = c("1-1-1", "1-1-2", "1-3-5-6", "1-3-2", "1-3-5-7-9"))
```

- Using the argument `too_many`:
```{r}
debug <- df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "debug"
  )

debug |> filter(!x_ok)
```

- The `x_remainder` column also shows the excess pieces.  
- We can opt to drop the excess pieces or merge them into the last column:
```{r}
# dropping the excess pieces
df |> 
  separate_wider_delim(
    x, 
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "drop"
  )

# merging the excess into the last column
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("x", "y", "z"),
    too_many = "merge"
  )
```


## Working with individual letters within a string

### Length

- Getting the string length using `str_length`:
```{r}
str_length(c("a", "R for data science", NA))
```

- Using `str_length()` in dplyr:
```{r}
babynames |> 
  count(length = str_length(name), wt = n)
```

- Another example:
```{r}
babynames |> 
  filter(str_length(name) == 15) |> 
  count(name, wt = n, sort = TRUE)
```

### Subsetting

- Using `str_sub()`:
```{r}
x <- c("Apple", "Banana", "Pear")

str_sub(x, 1, 3)
```

- The length of the subset is equal to `end - start + 1`.  

- Similar to Python, we can use `-1` to refer to the last character, -2 to the second to last character, etc. (but this is not 0-indexed!) 
```{r}
str_sub(x, -3, -1)
```

- But unlike Python, subsetting won't fail here if the string is too short -- it will just return as much as possible. 
```{r}
str_sub("a", 1, 5)
```

- Using `str_sub()` in dplyr:
```{r}
babynames |> 
  mutate(
    first = str_sub(name, 1, 1),
    last = str_sub(name, -1, -1)
  )
```


## Non-English text

- Using `charToRaw()` to convert a character to its hexadecimal form (aka the ASCII encoding):
```{r}
charToRaw("Hadley")
```


- While ASCII coverage for symbols is limited, UTF-8 can encode just about every character, as well as other symbols, like smileys and other icons and shapes.

- UTF-8 can also "fail" when the input uses a different character encoding. Here is an example using `read_csv` from the **readr** package:
```r 
x1 <- "text\nEl Ni\xf1o was particularly bad this year"
read_csv(x1)$text

x2 <- "text\n\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"
read_csv(x2)$text
```

- We can specify the correct encoding via the `locale` argument.
```r
read_csv(x1, locale = locale(encoding = "Latin1"))$text
read_csv(x2, locale = locale(encoding = "Shift-JIS"))$text
```

### Letter variations

- These two letters might look the same:
```{r}
u <- c("\u00fc", "u\u0308")
str_view(u)
```

- But they differ in length:
```{r}
str_length(u)
```

- However, the string comparator function `str_equal()` from **stringr** still recognizes that both have the same appearance:
```{r}
u[[1]] == u[[2]]
str_equal(u[[1]], u[[2]])
```


### Locale-dependent functions

- To check which locale is supported in **stringr**, use this command: `stringi::stri_locale_list()`.

- Example of difference in language rules depending on locale:
```{r}
str_to_upper(c("i", "ı"))
str_to_upper(c("i", "ı"), locale = "tr")
```

- Example of difference in sorting rules depending on locale:
```{r}
str_sort(c("a", "c", "ch", "h", "z"))
str_sort(c("a", "c", "ch", "h", "z"), locale = "cs")
```

- The `arrange()` function in **dplyr** is also locale-dependent.




