---
title: 'Notes on Chapter 18: Missing Values'
author: "Norman Lim"
date: "2025-08-03"
output:
  pdf_document: default
  html_document: default
---

### Prerequisites

```{r}
library(tidyverse)
```

### Last observation carried forward

```{r}
treatment <- tribble(
  ~person,            ~treatment, ~response,
  "Derrick Whitmore", 1,          7,
  NA,                 2,          10,
  NA,                 3,          NA,
  "Katherine Burke",  1,          4
)
```

- Using `fill()` to fill missing values by carrying the last observation forward aka **locf**:
```{r}
treatment |> 
  fill(everything())
```


### Filling missing values using fixed values:

- Using `coalesce()`:
```{r}
x <- c(1, 4, 5, 7, NA)

coalesce(x, 0)
```

Sometimes, in the data, other values are used to represent missing values, like `99` or `-999`. We can replace them with `NA` using the `na_if()` function:
```{r}
x <- c(1, 4, 5, 7, -99)
na_if(x, -99)
```

### NaN
`NaN` generally behaves like `NA`:
```{r}
x <- c(NA, NaN)
x * 10
x == 1
is.na(x)
```

- In case you need to distinguis `NA`, from `NaN`, use `is.nan()`:
```{r}
is.nan(x)
```

`NaN`s are usually produced when you perform a mathematical operation that has an indeterminate result:
```{r}
0 / 0
0 * Inf
Inf - Inf
sqrt(-1)
```


## Implicit missing values

- Example: an entire row of data is missing.
```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

stocks
```

- In the dataset above, price in the fourth quarter of 2020 is explicitly missing because its value is `NA`
- The `price` for the first quarter of 2021 is implicitly missing because it does not appear in the dataset. 

In other words:

> An explicit missing value is the presence of an absence.
> An implicit missing value is the absence of a presence.


### Pivoting

- Pivoting can make implicit missings explicit and vice versa:
```{r}
stocks |> 
  pivot_wider(
    names_from = qtr,
    values_from = price
  )
```

### Complete

`tidyr::complete()` allows you to generate explicit missing values by providing a set of variables that define the combination of rows that should exist. Example:
```{r}
stocks |> 
  complete(year, qtr)
```

If the individual variables are not complete aka there is not enough example for `complete()` to draw from, you can specify the missing part:
```{r}
stocks |> 
  complete(year = 2019:2021, qtr)
```

### Joins

The function `dplyr::anit_join(x, y)` is a useful tool in revealing rows in `x` that don't have match in `y`. Example:
```{r}
library(nycflights13)

flights |> 
  distinct(faa = dest) |> 
  anti_join(airports)


flights |> 
  distinct(tailnum) |> 
  anti_join(planes)
```


### Exercises
1. Can you find any relationship between the carrier and the rows that appear to be missing from `planes`?


## Factors and empty groups

This can happen if the factors do not contain all the possible entries specified in the labels. Example:
```{r}
health <- tibble(
  name = c("Ikaia", "Oletta", "Leriah", "Dahsay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age = c(34, 88, 75, 47, 56)
)

health |> count(smoker)
```

We can force `count()` to keep all the groups using `.drop = FALSE`:
```{r}
health |> count(smoker, .drop = FALSE)
```

Similar issues can occur when plotting:
```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete()

ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

Similar issues can occur when grouping. Again the solution is to supply the `.drop = FALSE` argument:
```{r}
health |> 
  group_by(smoker, .drop = FALSE) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  )
```

We got warnings and "interesting" results when we summarized the empty group. Empty vectors have `0` length, while missing values have length of `1`.

To avoid these results (and the warnings!), we can use `complete()` to make the implicit missing values explicit:
```{r}
health |> 
  group_by(smoker) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  ) |> 
  complete(smoker)
```

The main drawback of this method: the results of `count` was also rendered `NA` instead of 0 for the missing group. 

