---
title: "Notes on Ch24-Web Scraping"
author: "Norman Lim"
date: "2025-08-15"
output:
  pdf_document: default
  html_document: default
---

Main Idea: Introduce the basics of web scraping using **rvest**.

### Prerequisites
```{r}
library(tidyverse)
library(rvest)
```

## Scraping ethics and legalities

General principles:  

- Legalities depend a lot on where you live.  
- If data is public, factual, and non-personal, it is likely ok.  
- If data is not public, non-personal, or not factual, talk to a lawyer first.  
- If scraping many pages, make sure to wait a little more between each request.  
- If available, check the terms and service of the website.  
- Be careful when scraping personally identifiable information like names, email addresses, phone numbers, dates of birth, etc.  
- Look at copyright laws and see what's protected. They generally apply to original works of authorship, but scraping "facts" are usually excluded, but this may vary depending on location/country.  
- One may scrape data under the doctrine of fair use, but there is no hard and fast rule about fair use.  


## HTML basics

HTML has a hierarchical structure formed by **elements** which consist of tags and optional **attributes**, contents, and end tags.  

A good reference on html tags is the MDN Web Docs created by the team behind mozilla(?).


## Extracting data

Reading an HTML file using `read_html()`:
```{r}
html <- read_html("http://rvest.org/")
html
```


Writing a simple html page using `minimal_html()`:
```{r}
html <- minimal_html("
  <p>This is a paragraph</p>
  <ul>
    <li>This is a bulleted list</li>
  <>
")
html
```


### Find elements

CSS selector codes:  
- `p` selects all `<p>` elements.  
- `.title` selects all elements with class "title".  
- `#title` selects all elements with the `id` attribute that equals "title". `Id` attributes must be unique within a document, so this will only ever select a single element.  

Example: 
```{r}
html <- minimal_html("
  <h1>This is a heading</h1>
  <p id='first'>This is a parfagraph</p>
  <p class='important'>This is an important paragraph</p>
")
```

Using `html_elements()` to find all elements that match the selector:
```{r}
html |> html_elements("p")
html |> html_elements(".important")
html |> html_elements("#first")
```

The other function `html_element()` always returns the same number of outputs as inputs: 
```{r}
html |> html_element("p")
```

When you use a selector that doesn't match any elements, `html_elements()` returns a vector of length 0, while `html_element()` returns a missing value:
```{r}
html |> html_elements("b")
html |> html_element("b")
```


### Nesting selections

Typically, `html_elements()` is used to
identify elements that will become observations, and `html_element()` is used to find elements that will become variables. 
Example:

Creating the sample html document: 
```{r}
html <- minimal_html("
  <ul>
    <li><b>C-3PO</b> is a <i>droid</i> that weighs <span class='weight'>167 kg</span></li>
    <li><b>R4-P17</b> is a <i>droid</i></li>
    <li><b>R2-D2</b> is a <i>droid</i> that weighs <span class='weight'>96 kg</span></li>
    <li><b>Yoda</b> weighs <span class='weight'>66 kg</span></li>
  </ul>
  ")
```

Using `html_elements()` to make a vector where each element corresponds to a different character:  
```{r}
characters <- html |>  html_elements("li")
characters
```

Extracting the name of each character using `html_element()`: 
```{r}
characters |> html_element("b")
```

Getting the weigh value for each element: 
```{r}
characters |> html_element(".weight")
```

We got an `NA` since there are only three weight elements that are children `characters`. Using `html_elements()`: 
```{r}
characters |> html_elements(".weight")
```


### Text and attributes 

Extracting plain text contents of an html element using `html_text2()`: 
```{r}
characters |> 
  html_element("b") |> 
  html_text2()

characters |> 
  html_element(".weight") |> 
  html_text2()
```

The **rvest** package automatically handles escape characters. 

Extracting data from attributes using `html_attr()`:
```{r}
html <- minimal_html("
  <p><a href='https://en.wikipedia.org/wiki/Cat'>cats</a></p>
  <p><a href='https://en.wikipedia.org/wiki/Dog'>dogs</a></p>
")

html |> 
  html_elements("p") |> 
  html_elements("a") |> 
  html_attr("href")
```



### Tables

HTML tables are built up from four main elements:  
- `<table>`: the table itself.  
- `<tr>`: table row.  
- `<th>`: table heading.  
- `<td>`: table data.  

Sample html with table: 
```{r}
html <- minimal_html("
  <table class='mytable'>
  <tr><th>x</th>   <th>y</th></tr>
  <tr><td>1.5</td> <td>2.7</td></tr>
  <tr><td>4.9</td> <td>1.3</td></tr>
  <tr><td>7.2</td> <td>8.1</td></tr>
  </table>
")
```

Extracting table data using `html_table()`:
```{r}
html |> 
  html_element(".mytable") |> 
  html_table()
```

The values are automatically converted to numbers.


## Finding the right selectors

- Often, some experimenting is needed to find a selector that is both specific (doesn't select everything you don't care about) and sensitive (does select everything you care about).  
- the tool SelectorGadget is useful here.  
- The web toolkit for developers provided by Chrome is one of the best.  
- The CSS dinner tutorial on CSS selectors is highly recommended.  


## Putting it all together 


### StarWars

Structure of the target web page:  
```html
<section>
  <h2 data-id="1">The Phantom Menace</h2>
  <p>Released: 1999-05-19</p>
  <p>Director: <span class="director">George Lucas</span></p>
  
  <div class="crawl">
    <p>...</p>
    <p>...</p>
    <p>...</p>
  </div>
</section>
```

Goal: turn the data on the web page into a 7-row data frame with variables `title`, `year`, `director`, and `intro`. 

Reading the HTML and extracting all the `<section>` elements: 
```{r}
url <- "https://rvest.tidyverse.org/articles/starwars.html"
html <- read_html(url)

section <- html |> html_elements("section")
section
```


The code above retrieves seven elements matching the seven movies found on the web page, suggesting that using `section` as a selector is good.  

Extracting the individual elements:  
```{r}
section |> html_element("h2") |> html_text2()

section |> html_element(".director") |> html_text2()
```

Once we are sure that the results are what we want, we can wrap them up into a tibble: 
```{r}
tibble(
  title = section |> 
    html_element("h2") |> 
    html_text2(),
  released = section |> 
    html_element("p") |> 
    html_text2() |> 
    str_remove("Released: ") |> 
    parse_date(),
  director = section |> 
    html_element(".director") |> 
    html_text2(),
  intro = section |> 
    html_element(".crawl") |> 
    html_text2()
)
```


### IMDB top films 

Using `html_table()` to read the IMDB web page: 
```{r}
url <- "https://web.archive.org/web/20220201012049/https://www.imdb.com/chart/top/"

html <- read_html(url)

table <- html |> 
  html_element("table") |> 
  html_table()
table
```

Processung further: 
```{r}
ratings <- table |> 
  select(
    rank_title_year = `Rank & Title`,
    rating = `IMDb Rating`
  ) |> 
  mutate(
    rank_title_year = str_replace_all(rank_title_year, "\n +", " ")
  ) |> 
  separate_wider_regex(
    rank_title_year,
    patterns = c(
      rank = "\\d+", "\\. ",
      title = ".+", " +\\(",
      year = "\\d+", "\\)"
    )
  )
ratings
```


"Spelunking" the source web page for additional info:  
```{r}
html |> 
  html_elements("td strong") |> 
  head() |> 
  html_attr("title")
```

Combining this result with the previous result:  
```{r}
ratings |> 
  mutate(
    rating_n = html |> html_elements("td strong") |> html_attr("title")
  ) |> 
  separate_wider_regex(
    rating_n,
    patterns = c(
      "[0-9.]+ based on ",
      number = "[0-9,]+",
      " user ratings"
    )
  ) |> 
  mutate(
    number = parse_number(number)
  )
```


## Dynamic sites

At the time of writing, this area is still being worked on. The **chromote** package is used to give the user additional tools to interact with the target website.





