---
title: 'Notes on Chapter 23: Hierarchical Data'
author: "Norman Lim"
date: "2025-08-14"
output:
  pdf_document: default
  html_document: default
---

Main idea: This chapter is about **rectangling** -- the art of converting hierarchical data (or tree-like data) into rectangular data


### Prerequisites

Installing the required libraries:
```{r}
library(tidyverse)
library(repurrrsive)
library(jsonlite)
```


## Lists

Creating lists with `list()`:
```{r}
x1 <- list(1:4, "a", TRUE)

x1
```

Creating a list with named components or "children":
```{r}
x2 <- list(a = 1:2, b = 1:3, c = 1:4)

x2
```


Displaying the structure of a list with `str()`:
```{r}
str(x1)
str(x2)
```

Note: this takes up less space than printing the contents of a list. The result looks a bit like the output of `glimpse()` for dataframes.


### Hierarchy

Creating a nested list (a list within a list):
```{r}
x3 <- list(list(1, 2), list(3, 4))
str(x3)
```

The concatenator `c()`, on the other hand, creates a flat vector:
```{r}
c(c(1, 2), c(3, 4))
x4 <- c(list(1, 2), list(3, 4))
str(x4)
```

Example of an extremely-nested list:
```{r}
x5 <- list(1, list(2, list(3, list(4, list(5)))))
str(x5)
```

As the list gets more complex, it becomes more difficult to see its structure, even when `str()` is used. The command `View()` is more practical when working with complex lists. This command will show the contents of a list in a gui. 


### List-columns

Lists that live inside a tibble are called list-columns.

Example of a list-column:
```{r}
df <- tibble(
  x = 1:2,
  y = c("a", "b"),
  z = list(list(1, 2), list(3, 4, 5))
)
df
```

Using **dplyr** functions in a tibble with list-columns:
```{r}
df |> 
  filter(x == 1)
```


The list-columns behave like any other column, but they require more compute resources. The process of unnesting can help mitigate this drawback.


## Unnesting

Basic forms of list-columns:  

- **named:** The names are usually consistent for every row.  
- **unnamed:** The number of elements tends to vary for each row.  

Example of a named list-column:
```{r}
df1 <- tribble(
  ~x, ~y,
  1, list(a = 11, b = 12),
  2, list(a = 21, b = 22),
  3, list(a = 31, b = 32)
)
df1
```

Example of an unnamed list-column:
```{r}
df2 <- tribble(
  ~x, ~y,
  1, list(11, 12, 13),
  2, list(21),
  3, list(31, 32)
)
df2
```

For these two cases, the functions `unnest_wider()` and `unnest_longer()` are commonly used.

### `unnest_wider()`

This is commonly used when each row has the same number of elements, like in the case of a named list-column:
```{r}
df1 |> 
  unnest_wider(y)
```

The names of the new columns were, by default, obtained from the names of the list elements. 

Changing the default name for the new columns generated by `unnest_wider()` using the `names_sep` argument:
```{r}
df1 |> 
  unnest_wider(y, names_sep = "_")
```


### `unnest_longer()`

Unnesting a tibble with an unnamed list-column using `unnest_longer()`:
```{r}
df2 |> 
  unnest_longer(y)
```

In this case, the values for `x` column gets duplicated for each element inside `y`. In case one of the elements is empty, we get:
```{r}
df6 <- tribble(
  ~x, ~y,
  "a", list(1, 2),
  "b", list(3),
  "c", list()
)

df6 |> unnest_longer(y)
```

With nothing to "distribute" `x` with, the row disappears. If you want to preserve the row with missing list, the argument `keep_empty = TRUE` can be used:
```{r}
df6 |> unnest_longer(y, keep_empty = TRUE)
```


### Inconsistent types

Unnesting a data with a list-column that contains different types of vector:
```{r}
df4 <- tribble(
  ~x, ~y,
  "a", list(1),
  "b", list("a", TRUE, 5)
)

df4 |> unnest_longer(y)
```

In this case, `unnest_longer()` cannot fully unnest the list-column because it can't find a common type of vector. The original list-column will only be unnested until single-element lists remain, in order for the new column to have a consistent vector type -- a list, in this case.


### Other functions

- `unnest_auto()`: automatically pics between `unnest_longer()` and `unnest_wider()` depending on the structure of the list-column. It might be convenient to use initially but it can "spoil" the user because it doesn't force the user to understand how the data is structured.  
- `unnest()`: expands both rows and columns. Useful when the list-column contains a 2d structure. 


## Case studies

Let's begin by creating a very wide data:
```{r}
repos <- tibble(json = gh_repos)
repos
```

The tibble contains 6 rows, but each row contains an unnamed list with either 26 or 30 elements.   

Since these lists are unnamed, it can be a good idea to start with `unnest_longer()` to deal with the data:
```{r}
repos |> 
  unnest_longer(json)
```

Now each element is named so it is now possible to use `unnest_wider()`:
```{r}
repos |> 
  unnest_longer(json) |> 
  unnest_wider(json)
```

The operation above works! However, the resulting data has so many columns that it is till difficult to see even the column names.   

Displaying the first 10 column names:
```{r}
repos |> 
  unnest_longer(json) |> 
  unnest_wider(json) |> 
  names() |> 
  head(10)
```

Selecting only the relevant columns (or the "interesting" ones):
```{r}
repos |> 
  unnest_longer(json) |> 
  unnest_wider(json) |> 
  select(id, full_name, owner, description)
```

Now it is a bit easier to make sense of the data. Each child in the hierarchy is a GitHub user and the list for each user contains the GitHub repositories that they created.  

If we try expanding or unnesting the `owner` list-column, we would get an error message:
```r
repos |> 
  unnest_longer(json) |> 
  unnest_wider(json) |> 
  select(id, full_name, owner, description) |> 
  unnest_wider(owner)
#> Error in `unnest_wider()`:
#> ! Can't duplicate names between the affected columns and the original
#>   data.
#> x These names are duplicated:
#>   i `id`, from `owner`.
#> i Use `names_sep` to disambiguate using the column name.
#> i Or use `names_repair` to specify a repair strategy.
```

This is because the list-column `owner` also contains an `id` column and we already have a column with this name. This is where the `names_sep` argument comes in handy:
```{r}
repos |> 
  unnest_longer(json) |> 
  unnest_wider(json) |> 
  select(id, full_name, owner, description) |> 
  unnest_wider(owner, names_sep = "_")
```


### Relational data

Nested data is sometimes used to represent that is usually spread across multiple data frames. 
```{r}
chars <- tibble(json = got_chars)
chars
```

Again, this is has an unnamed list-column so it is a good idea to start with `unnest_wider()`:
```{r}
chars |> 
  unnest_wider(json)
```

Selecting "interesting" or relevant columns, to make the result easier to read:
```{r}
characters <- chars |> 
  unnest_wider(json) |> 
  select(id, name, gender, culture, born, died, alive)

characters
```

The columns chosen were not list-columns. Where are the list-columns? To find them, the functions `where()` and the `is.list` can be used:
```{r}
chars |> 
  unnest_wider(json) |> 
  select(id, where(is.list))
```

Exploring the `titles` column, which we can now see as an unnamed list-column:
```{r}
chars |> 
  unnest_wider(json) |> 
  select(id, titles) |> 
  unnest_longer(titles)
```

The result is a table. To flatten the data, we can use joins to merge the tables nested in each list-column. But first we need to clean them:
```{r}
titles <- chars |> 
  unnest_wider(json) |> 
  select(id, titles) |> 
  unnest_longer(titles) |> 
  filter(titles != "") |> 
  rename(title = titles)

titles
```

We can then proceed to create a table like this for each list-column in the data and then using joins to combine them with the `character` data to produce flattened version of the original.  


### Deeply nested

In this case, repeated unnesting is necessary.
```{r}
gmaps_cities
```

The data looks initially "harmless" -- just a two-column dataset, with the second column being a 2-element list-column. Unnesting, we get:
```{r}
gmaps_cities |> 
  unnest_wider(json)
```

To make things simpler in this example, we will drop the `status` column and continue to explore the remaining list-column:
```{r}
gmaps_cities |> 
  unnest_wider(json) |> 
  select(-status) |> 
  unnest_longer(results)
```

The result, again, is a named list, so we can use `unnest_wider()`:
```{r}
locations <- gmaps_cities |> 
  unnest_wider(json) |> 
  select(-status) |> 
  unnest_longer(results) |> 
  unnest_wider(results)

locations
```

Exploring the exact location of cities with similar names:
```{r}
locations |> 
  select(city, formatted_address, geometry) |> 
  unnest_wider(geometry)
```

Unnesting the list-column `location` will give us the coordinates:
```{r}
locations |> 
  select(city, formatted_address, geometry) |> 
  unnest_wider(geometry) |> 
  unnest_wider(location)
```

Extracting the `bounds`:
```{r}
locations |> 
  select(city, formatted_address, geometry) |> 
  unnest_wider(geometry) |> 
  select(!location:viewport) |> 
  unnest_wider(bounds)
```

Renaming the `southwest` and `northeast` corners of the rectangular bounds to make it shorter and more intuitive:
```{r}
locations |> 
  select(city, formatted_address, geometry) |> 
  unnest_wider(geometry) |> 
  select(!location:viewport) |> 
  unnest_wider(bounds) |> 
  rename(ne = northeast, sw = southwest) |> 
  unnest_wider(c(ne, sw), names_sep = "_")
```

Using the function `hoist()` to extract directly the deeply-nested elements that we are interested in (similar results to the multiple unnesting above):
```{r}
locations |> 
  select(city, formatted_address, geometry) |> 
  hoist(
    geometry,
    ne_lat = c("bounds", "northeast", "lat"),
    sw_lat = c("bounds", "southwest", "lat"),
    ne_lng = c("bounds", "northeast", "lng"),
    sw_lng = c("bounds", "northeast", "lng")
  )
```

This is neat way of dealing with deeply-nested data, but we can only use this after understanding the structure of our dataset. If a data dictionary is provided, we might be able to use this with less data exploration. 


## JSON

This file structure is designed to be easily read (and written) by machines. Generally has six data types:  
- the null type: like `NA` in R.  
- the string type: always wrapped in double quotes.  
- the number type: can be integers, decimals, and can be in scientific notation, but does not support `Inf` or `NaN` values.  
- the boolean type: logical conditions in lower case (`true` and `false`).  

An **unnamed** list is called an **array** in JSON.   
A **named** list is called an **object**.


### jsonlite

Using the `jsonlite` package to convert JSON file into one with R data structure:
```{r}
# a path to a json file inside the jsonlite package
gh_users_json()

# reading
gh_users2 <- read_json(gh_users_json())

# Checking
identical(gh_users, gh_users2)
```

Parsing strings with json structure:
```{r}
str(parse_json('1'))
str(parse_json('[1, 2, 3]'))
str(parse_json('{"x": [1, 2, 3]}'))
```


### Starting the rectangling process

Creating a sample tibble with JSON structure:
```{r}
json <- '[
  {"name": "John", "age": 34},
  {"name": "Susan", "age": 27}
]'

df <- tibble(json = parse_json(json))
df
```

Unnesting the sample data:
```{r}
df |> 
  unnest_wider(json)
```

A JSON file with a single top-level JSON object:
```{r}
json <- '{
  "status": "OK",
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
  ]
}'
```

To unnest this, we need to first wrap it in a list before we put it in a tibble:
```{r}
df <- tibble(json = list(parse_json(json)))
df
```

Unnesting:
```{r}
df |> 
  unnest_wider(json) |> 
  unnest_longer(results) |> 
  unnest_wider(results)
```

An alternative approach is to reach inside the parsed JSON and start with the part that we are interested in:
```{r}
df <- tibble(results = parse_json(json)$results)

df |> 
  unnest_wider(results)
```




